---
title: RFC - Enhancing tapply
author: Deepayan Sarkar
---


```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, autodep = TRUE,
               comment = "", warning = TRUE, message = FALSE, prompt = FALSE)
options(warnPartialMatchDollar = FALSE, width = 100)
suppressMessages(library(dplyr))
```


# Motivation

For group-wise summaries of a data frame, `dplyr` can do the following.

- Several scalar-valued functions

```{r}
ToothGrowth |> group_by(supp, dose) |>
    summarize(m = mean(len), s = sd(len), n = length(len))
```

- Scalar valued functions with vector input

```{r}
ToothGrowth |> group_by(supp) |>
    summarize(mlen = mean(len), mdose = mean(dose),
              slope = coef(lm(len ~ log(dose)))[[2]])
```

- Vector valued functions with scalar / vector input

```{r}
ToothGrowth |> group_by(supp) |>
    summarize(data.frame(mlen = mean(len), slen = sd(dose)),
              as.data.frame(as.list(coef(lm(len ~ log(dose))))))
```

But this means that we have to have convoluted code for something like

```{r}
ToothGrowth |> group_by(supp, dose) |>
    summarize(data.frame(as.list(quantile(len))))
```

because the simpler option doesn't work.

```{r}
ToothGrowth |> group_by(supp, dose) |> summarize(quantile(len))
```

Another interesting example is where the input data frame is
summarized in some way, but not to scalars. Let's say order by first
column and then take first 3 rows.


```{r}
mySummary <- function(d)
{
    ## could also use dplyr::rearrange() instead
    o <- order(d[[1]])
    head(d[o, ], 3)
}
```


```{r}
ToothGrowth |> group_by(supp) |>
    summarize(mlen = mean(len), 
              mySummary(data.frame(len = len, dose = dose)))
```

This was OK with `dplyr_1.0.2`, but deprecated in `dplyr_1.1.0`.

What are closest equivalents with base R?

# split

Very general, but needs manual post-processing

```{r}
s1 <- split(ToothGrowth, ~ supp + dose)
lapply(s1, with, data.frame(supp = unique(supp),
                            dose = unique(dose),
                            m = mean(len),
                            s = sd(len),
                            n = length(len))) |>
    do.call(what = rbind)

s2 <- split(ToothGrowth, ~ supp)
lapply(s2, with, data.frame(supp = unique(supp),
                            mlen = mean(len),
                            mdose = mean(dose),
                            slope = coef(lm(len ~ log(dose)))[[2]])) |>
    do.call(what = rbind)

lapply(s2, with, data.frame(supp = unique(supp),
                            mlen = mean(len),
                            slen = sd(dose),
                            as.list(coef(lm(len ~ log(dose)))))) |>
    do.call(what = rbind)

lapply(s2, with, data.frame(supp = unique(supp),
                            mlen = mean(len),
                            mySummary(data.frame(len = len, dose = dose)))) |>
    do.call(what = rbind)
```



# aggregate

- Several scalar-valued functions

Can only do one at a time (and merge 2 at a time). Not ideal.

```{r}
merge(aggregate(len ~ supp + dose, ToothGrowth, mean),
      aggregate(len ~ supp + dose, ToothGrowth, sd),
      by = c("supp", "dose"))

## can't name unless more than one result
aggregate(len ~ supp + dose, ToothGrowth,
          function(x) c(m = mean(x), foo = 1))
```


- Scalar valued functions with vector input: not possible using `aggregate()`

- Vector valued functions with scalar / vector input

Scalar input OK, vector not possible.

```{r}
aggregate(len ~ supp, ToothGrowth,
          function(x) c(m = mean(x), s = sd(x)))
```

On the other hand, can do nice things for scalar input like

```{r}
aggregate(len ~ supp + dose, ToothGrowth, quantile)
```

One limitation is that `aggregate()` doesn't seem to be able to handle
non-atomic vector output; e.g.,

```{r}
aggregate(len ~ supp, ToothGrowth,
          function(x) data.frame(m = mean(x), s = sd(x)))
```


# tapply

Currently doesn't work for data frame, but with check modified:

```
     nI <- length(INDEX)  # now, 'INDEX' is not classed
     if (!nI) stop("'INDEX' is of length zero")
-    if (!all(lengths(INDEX) == length(X)))
+    if (!is.object(X) && !all(lengths(INDEX) == length(X)))
         stop("arguments must have same length")
     namelist <- lapply(INDEX, levels)#- all of them, yes !
     extent <- lengths(namelist, use.names = FALSE)
     cumextent <- cumprod(extent)
```


```{r}
tapply(ToothGrowth, with(ToothGrowth, list(supp)),
       function(d) with(d, data.frame(supp = unique(supp),
                                      mlen = mean(len),
                                      as.list(coef(lm(len ~ log(dose)))))))

s <- 
tapply(ToothGrowth, with(ToothGrowth, list(supp, dose)),
       function(d) with(d, data.frame(supp = unique(supp),
                                      dose = unique(dose),
                                      mlen = mean(len),
                                      mdose = mean(log(dose)))))

do.call(rbind, s)
```

We could actually avoid the function definition and use NSE instead if we use `FUN = with`.

```{r}
s <- 
    tapply(ToothGrowth, with(ToothGrowth, list(supp, dose)),
           with, data.frame(supp = unique(supp),
                            dose = unique(dose),
                            mlen = mean(len),
                            mdose = mean(log(dose))))
do.call(rbind, s)
```

This also works with multi-row functions.

```{r}
tapply(ToothGrowth, ToothGrowth$supp,
       with, data.frame(supp = unique(supp),
                        mySummary(data.frame(len = len, dose = dose)))) |>
    do.call(what = rbind)
```


This is already quite nice, except that we need to include the unique
values of the splitting factors in the summary function.


# Possible extensions

## `aggregate`

We could extend `aggregate` so that

```r
aggregate( ~ supp + dose, ToothGrowth, FUN)
```

applies `FUN` on the components of `split(ToothGrowth, ~ supp +
dose)`. This is currently an error, so should be OK.

But I am not sure when and how to allow arbitrary functions that
operate on the sub-dataframes.


## `tapply`

`tapply` has more immediate promise. We can support formulas fairly
easily, similar to split:

```
tapply <- function (X, INDEX, FUN = NULL, ..., default = NA, simplify = TRUE)
 {
     FUN <- if (!is.null(FUN)) match.fun(FUN)
+    if (inherits(INDEX, "formula")) INDEX <- formula2varlist(INDEX, X)
     if (!is.list(INDEX)) INDEX <- list(INDEX)
```

with a new `formula2varlist()`, a version of which is already used in
`split()` without the checks:

```r
formula2varlist <- function(formula, data, warnLHS = TRUE, ignoreLHS = warnLHS)
{
    if (!inherits(formula, "formula")) stop("'formula' must be a formula")
    if (length(formula) == 3) {
        if (warnLHS) {
            if (ignoreLHS)
                warning("Unexpected LHS in 'formula' has been ignored.")
            else
                warning("Unexpected LHS in 'formula' has been combined with RHS.")
        }
        if (ignoreLHS) formula <- formula[-2]
    }
    fterms <- stats::terms(formula)
    ans <- eval(attr(fterms, "variables"), data, environment(formula))
    names(ans) <- attr(fterms, "term.labels")
    ans
}
```

With this, we have

```{r}
tapply(ToothGrowth, ~ supp + dose, with, mean(len))
tt <- tapply(ToothGrowth, ~ supp + dose,
             with, data.frame(m = mean(len), s = sd(len)),
             simplify = FALSE)
tt
tt[2,2][[1]]
do.call(rbind, tt)
```

All we need now is a enhanced version of rbind which adds the margins
as columns, as `as.data.frame.table()` does.

## Reversing `tapply()`

We basically want something that reverses `tapply()`. For scalar
output and a single factor, `lattice::make.groups()` works.

```{r}
tapply(ToothGrowth, ~ supp, with, mean(len), simplify = FALSE) |>
    do.call(what = lattice::make.groups)
```

Should we think of something like that but more general?

