---
title: RFC - Enhancing tapply
author: Deepayan Sarkar
---


```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, autodep = TRUE,
               comment = "", warning = TRUE, message = FALSE, prompt = FALSE)
options(warnPartialMatchDollar = FALSE, width = 100)
suppressMessages(library(dplyr))
```


# Motivation

For group-wise summaries of a data frame, `dplyr` can do the following.

- Several scalar-valued functions

```{r}
ToothGrowth |> group_by(supp, dose) |>
    summarize(m = mean(len), s = sd(len), n = length(len))
```

- Scalar valued functions with vector input

```{r}
ToothGrowth |> group_by(supp) |>
    summarize(mlen = mean(len), mdose = mean(dose),
              slope = coef(lm(len ~ log(dose)))[[2]])
```

- Vector valued functions with scalar / vector input

```{r}
ToothGrowth |> group_by(supp) |>
    summarize(data.frame(mlen = mean(len), slen = sd(dose)),
              as.data.frame(as.list(coef(lm(len ~ log(dose))))))
```

But this means that we have to have convoluted code for something like

```{r}
ToothGrowth |> group_by(supp, dose) |>
    summarize(data.frame(as.list(quantile(len))))
```

because the simpler option doesn't work.

```{r}
ToothGrowth |> group_by(supp, dose) |> summarize(quantile(len))
```

Another interesting example is where the input data frame is
summarized in some way, but not to scalars. Let's say order by first
column and then take first 3 rows.


```{r}
mySummary <- function(d)
{
    ## could also use dplyr::rearrange() instead
    o <- order(d[[1]])
    head(d[o, ], 3)
}
```


```{r}
ToothGrowth |> group_by(supp) |>
    summarize(mlen = mean(len), 
              mySummary(data.frame(len = len, dose = dose)))
```

This was OK with `dplyr_1.0.2`, but deprecated in `dplyr_1.1.0`.

What are closest equivalents with base R?

# split

Very general, but needs manual post-processing

```{r}
s1 <- split(ToothGrowth, ~ supp + dose)
lapply(s1, with, data.frame(supp = unique(supp),
                            dose = unique(dose),
                            m = mean(len),
                            s = sd(len),
                            n = length(len))) |>
    do.call(what = rbind)

s2 <- split(ToothGrowth, ~ supp)
lapply(s2, with, data.frame(supp = unique(supp),
                            mlen = mean(len),
                            mdose = mean(dose),
                            slope = coef(lm(len ~ log(dose)))[[2]])) |>
    do.call(what = rbind)

lapply(s2, with, data.frame(supp = unique(supp),
                            mlen = mean(len),
                            slen = sd(dose),
                            as.list(coef(lm(len ~ log(dose)))))) |>
    do.call(what = rbind)

lapply(s2, with, data.frame(supp = unique(supp),
                            mlen = mean(len),
                            mySummary(data.frame(len = len, dose = dose)))) |>
    do.call(what = rbind)
```



# aggregate

- Several scalar-valued functions

Can only do one at a time (and merge 2 at a time). Not ideal.

```{r}
merge(aggregate(len ~ supp + dose, ToothGrowth, mean),
      aggregate(len ~ supp + dose, ToothGrowth, sd),
      by = c("supp", "dose"))

## can't name unless more than one result
aggregate(len ~ supp + dose, ToothGrowth,
          function(x) c(m = mean(x), foo = 1))
```


- Scalar valued functions with vector input: not possible using `aggregate()`

- Vector valued functions with scalar / vector input

Scalar input OK, vector not possible.

```{r}
aggregate(len ~ supp, ToothGrowth,
          function(x) c(m = mean(x), s = sd(x)))
```

On the other hand, can do nice things for scalar input like

```{r}
aggregate(len ~ supp + dose, ToothGrowth, quantile)
```

One limitation is that `aggregate()` doesn't seem to be able to handle
non-atomic vector output; e.g.,

```{r}
aggregate(len ~ supp, ToothGrowth,
          function(x) data.frame(m = mean(x), s = sd(x)))
```


# tapply

From R 4.3.0, `tapply()` can handle data frame input. `tapply()` does
not do NSE, but we can avoid the function form and use NSE if we use
`FUN = with`. For example,

```{r}
tvec <- 
    tapply(ToothGrowth, ~ supp, with,
           c(supp = unique(supp),
             mlen = mean(len),
             coef(lm(len ~ log(dose)))))
tdf <- 
    tapply(ToothGrowth, ~ supp, with,
           data.frame(supp = unique(supp),
                      mlen = mean(len),
                      what = c("int", "slope"),
                      coefs = coef(lm(len ~ log(dose)))))
```

The second one returns multi-row data frames.

The result is usually a list matrix. 

```{r}
str(tdf)
dimnames(tdf)
```



In both the examples above, `rbind()` will give a reasonable matrix or
data frame that one can work with.

```{r}
do.call(rbind, tvec)
do.call(rbind, tdf)
```

This is already quite nice, except that we need to include the unique
values of the splitting factors in the summary function. Usually one
will not do so, and a conversion to data frame should include those.

```{r}
tvec0 <- 
    tapply(ToothGrowth, ~ supp, with,
           c(mlen = mean(len),
             coef(lm(len ~ log(dose)))))
tdf0 <- 
    tapply(ToothGrowth, ~ supp, with,
           data.frame(mlen = mean(len),
                      what = c("int", "slope"),
                      coefs = coef(lm(len ~ log(dose)))))
## conventional example
qvec0 <- with(ToothGrowth,
              tapply(len, list(s = supp, d = dose),
                     quantile, prob = c(1/4, 2/4, 3/4)))
do.call(rbind, tdf0)
```


The following function proposes to do this.


```{r}
array2DF <- function (x, responseName = "Value", sep = "",
                      base = list(LETTERS), simplify = FALSE)
{
    keys <-
        do.call("expand.grid",
                c(dimnames(provideDimnames(x, sep = sep, base = base)),
                  KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE))
    vals <- NULL
    if(simplify) {
        dfrows <- .compatibleDFs(x)
        if (length(dfrows))
            return(cbind(keys[ rep(seq_along(dfrows), dfrows), , drop = FALSE],
                         do.call(rbind, x)))
        x <- simplify2array(c(x))
        str(x)
        if(is.array(x)) {
            print(sapply(x, length))
            vals <- asplit(x, 1L)
            if(is.null(names(vals)))
                names(vals) <-
                    paste0(responseName, sep, seq_along(vals))
        }
    }
    if(is.null(vals)) {
        vals <- list(c(x))
        names(vals) <- responseName
    }
    cbind(keys, list2DF(vals))
}


.compatibleDFs <- function(x)
{
    ## check whether all components of list array 'x' 
    ## - are data frames
    ## - have same column names
    ## If TRUE, return value is a vector of corresponding nrow()-s 
    ## If FALSE, return value is integer(0)
    if (!all(sapply(x, inherits, "data.frame"))) return(integer(0))
    if (length(unique(sapply(x, ncol))) > 1L) return(integer(0))
    if (length(unique(lapply(x, colnames))) > 1L) return(integer(0))
    return(sapply(x, nrow))
}
    
```

In action: without simplification.

```{r}
array2DF(tvec0, simplify = FALSE) |> str()
array2DF(tdf0, simplify = FALSE) |> str()
array2DF(qvec0, simplify = FALSE) |> str()
```

With simplification.

```{r}
## OK except for unnecessary dimnames
array2DF(qvec0, simplify = TRUE) |> str()
array2DF(tvec0, simplify = TRUE) |> str()

array2DF(tdf0, simplify = TRUE)
```

What about unnamed results? Say we want within group bootstrap.

Make unbalanced to see how that affects result.

```{r}
b <- 
    tapply(ToothGrowth[-1,], ~ supp + dose, with,
           sample(len, replace = TRUE))
array2DF(b, simplify = TRUE)
```

I think it would be better to `c()` here.


Something silly: 

```{r}
x <- 
    tapply(ToothGrowth, ~ supp + dose, with,
           if (runif(1) < 0.5) mean(len) else quantile(len))
array2DF(x, simplify = TRUE)
```

Looks OK.


# Possible extensions

## `aggregate`

We could extend `aggregate` so that

```r
aggregate( ~ supp + dose, ToothGrowth, FUN)
```

applies `FUN` on the components of `split(ToothGrowth, ~ supp +
dose)`. This is currently an error, so should be OK.

But I am not sure when and how to allow arbitrary functions that
operate on the sub-dataframes.


## `tapply`

`tapply` has more immediate promise. We can support formulas fairly
easily, similar to split:

```
tapply <- function (X, INDEX, FUN = NULL, ..., default = NA, simplify = TRUE)
 {
     FUN <- if (!is.null(FUN)) match.fun(FUN)
+    if (inherits(INDEX, "formula")) INDEX <- formula2varlist(INDEX, X)
     if (!is.list(INDEX)) INDEX <- list(INDEX)
```

with a new `formula2varlist()`, a version of which is already used in
`split()` without the checks:

```r
formula2varlist <- function(formula, data, warnLHS = TRUE, ignoreLHS = warnLHS)
{
    if (!inherits(formula, "formula")) stop("'formula' must be a formula")
    if (length(formula) == 3) {
        if (warnLHS) {
            if (ignoreLHS)
                warning("Unexpected LHS in 'formula' has been ignored.")
            else
                warning("Unexpected LHS in 'formula' has been combined with RHS.")
        }
        if (ignoreLHS) formula <- formula[-2]
    }
    fterms <- stats::terms(formula)
    ans <- eval(attr(fterms, "variables"), data, environment(formula))
    names(ans) <- attr(fterms, "term.labels")
    ans
}
```

With this, we have

```{r}
tapply(ToothGrowth, ~ supp + dose, with, mean(len))
tt <- tapply(ToothGrowth, ~ supp + dose,
             with, data.frame(m = mean(len), s = sd(len)),
             simplify = FALSE)
tt
tt[2,2][[1]]
do.call(rbind, tt)
```

All we need now is a enhanced version of rbind which adds the margins
as columns, as `as.data.frame.table()` does.

## Reversing `tapply()`

We basically want something that reverses `tapply()`. For scalar
output and a single factor, `lattice::make.groups()` works.

```{r}
tapply(ToothGrowth, ~ supp, with, mean(len), simplify = FALSE) |>
    do.call(what = lattice::make.groups)
```

Should we think of something like that but more general?

