---
layout: remark
title: Measuring Current Growth Rate of COVID Infections
author: Deepayan Sarkar
mathjax: true
---


layout: true

# Background

---

<div>
$$
\newcommand{\sub}{_}
$$
</div>

* Trying to measure growth rate rather than absolute number of cases

* This is related to [this ad hoc measure](increment-ratio.html), but more grounded in theory

* Motivation: 

	- Let $x(t)$ be total number infected at time $t$
	
	- Let $a(t)$ be number of active infections at time $t$
	
	- Let $\lambda(t)$ be the _number of new infections per active infection per unit time_ at time $t$
	
	- Then a reasonable way to model the evolution of $x(t)$ is
	$$
	x(t + \delta t) - x(t) \approx \delta t \, a(t) \, \lambda(t)
	$$

---

- This gives an instantaneous definition of $\lambda(t)$ as
$$
\lambda(t) = \frac{ \frac{dx(t)}{dt} }{ a(t) } \approx \frac{ \delta x(t) }{ \delta t \,  a(t) }
$$

- We do have daily data, but using $\delta t = 1$ without smoothing gives very noisy results

- We instead use  $\delta t = 7$

- In theory, this measure can be compared across time and location
	
	- It should not depend on degree of systematic under-reporting

	- It should not be overly affected by daily fluctuations and patterns

- We report $R(t) := 10 * \lambda(t)$ assuming each active person is infectious for 10 days


- Any $R(t)$ higher than 1 is bad on a sustained basis

- What is a threhold for it to be _really_ bad?

---

- Calculations based on data from [covid19india.org](https://covid19india.org)

- $x(t) = \text{ Confirmed}$

- $a(t) = x(t) - \text{ Recovered } - \text{ Deceased }$


```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, autodep = TRUE,
               comment = "", warning = TRUE, message = FALSE, prompt = FALSE,
               dev.args = list(pointsize = 16), echo = FALSE, eval = TRUE,
               fig.width = 15, fig.height = 8, dpi = 96, fig.path='figures/growth-rate-')
options(warnPartialMatchDollar = FALSE, width = 100)
library(lattice)
library(RColorBrewer)
library(latticeExtra)
bdark <- RColorBrewer::brewer.pal(n = 8, name = "Dark2")
ct <- custom.theme(symbol = bdark, fill = adjustcolor(bdark, alpha.f = 0.5))
ct$strip.background$col <- "grey90"
ct$strip.border$col <- "grey50"
ct$axis.line$col <- "grey50"
lattice.options(default.theme = ct, default.args = list(as.table = TRUE))
```




```{r}
source("smoothing-experiments/tregularize.R")
source("covid-utils.R")
TARGET <- "districts.csv"
## To download the latest version, delete the file and run again
if (!file.exists(TARGET))
    download.file("https://api.covid19india.org/csv/latest/districts.csv",
                  destfile = TARGET)
dall <- read.csv(TARGET)
dall <-
    within(dall,
    {
        Date  <- as.Date(Date)
    })
LAG <- 7
```

.center[

[ Updated: `r {max(dall$Date)}` ]

]


---

layout: false

# Cities: Bengaluru, Chennai, Delhi, Kolkata, Mumbai


```{r}
LAMBDA <- 0
rbind(getRatioActive(dall, district = "Delhi", lag = LAG),
      getRatioActive(dall, district = "Mumbai", lag = LAG),
      getRatioActive(dall, district = "Chennai", lag = LAG),
      getRatioActive(dall, district = "Bengaluru Urban", lag = LAG),
      getRatioActive(dall, district = "Kolkata", lag = LAG)) |>
         xyplot(x = rcases ~ date | where, subset = date > as.Date("2020-07-01"),
                ylab = "Ratio of last two 1-week increments in confirmed cases",
                ## ylim = extendrange(c(0, 3)),
                prepanel = prepanelq,
                abline = list(h = 1, lwd = 2, col = "grey"),
                type = "l", pch = 16, grid = TRUE)
```

---

# States: Delhi, Karnataka, Maharashtra, West Bengal

```{r}
LAMBDA <- FALSE
rbind(getRatioActive(dall, "Delhi", lag = LAG, smooth = LAMBDA),
      getRatioActive(dall, "Maharashtra", lag = LAG, smooth = LAMBDA),
      getRatioActive(dall, "Karnataka", lag = LAG, smooth = LAMBDA),
      getRatioActive(dall, "West Bengal", lag = LAG, smooth = LAMBDA)) |>
         xyplot(x = rcases ~ date | where, subset = date > as.Date("2020-07-01"),
                ylab = "Ratio of last two 1-week increments in confirmed cases",
                ## ylim = extendrange(c(0, 3)),
                prepanel = prepanelq,
                layout = c(NA, 2),
                abline = list(h = 1, lwd = 2, col = "grey"),
                type = "l", pch = 16, grid = TRUE)
```


---

# Current growth rates by state

.scrollable500[

```{r}
LAMBDA <- FALSE
latestRatio <- function(s, min.cases = 100)
{
    r <- getRatioActive(dall, state = s, lag = LAG, smooth = LAMBDA)
    c(round(rev(r$rcases)[c(1, 8)], 2),
      rev(r$icases)[c(1, 8)])
}
states <- unique(dall$State)
rstates <- as.data.frame(t(sapply(states, latestRatio)))
names(rstates) <-
    c("Current Rate", "Rate One week ago",
      "Cases in last 7 days", "Cases in previous 7 days")
kable(rstates[ order(rstates[[3]], decreasing = TRUE), ])
```

]


```{r}
states <- rownames(rstates)[is.finite(rstates[[1]])]
LAMBDA <- FALSE
slist <- lapply(sort(states),
                function(s) getRatioActive(dall, s, lag = LAG, smooth = LAMBDA, death.lag = 7))
sdf <- do.call(rbind, slist)
p <- xyplot(x = rcases ~ date | where, data = sdf,
            subset = (total >= 1e5 & date > as.Date("2020-07-01") & 
                      !(where %in% c("Delhi", "Maharashtra", "Karnataka", "West Bengal"))),
            ylab = "Ratio of last two 1-week increments in confirmed cases",
            ## ylim = extendrange(c(0, 3)),
            prepanel = prepanelq,
            abline = list(h = 1, lwd = 2, col = "grey"),
            type = "l", pch = 16, grid = TRUE)
N <- dim(p)
```

---

layout: true

# All other states with at least 100000 confirmed cases

---

```{r}
if (N > 6) p[1:6] else p
```

---

```{r}
if (N > 6) {
    if (N > 12) p[7:12] else p[7:N]
}
```

---

```{r}
if (N > 12) {
    if (N > 18) p[13:18] else p[13:N]
}
```

```{r}
if (N > 18) {
    p[19:N]
}
```

---

layout: false

# All election states have seen a surge

```{r}
LAMBDA <- FALSE
rbind(getRatioActive(dall, "Assam", lag = LAG, smooth = LAMBDA),
      getRatioActive(dall, "Kerala", lag = LAG, smooth = LAMBDA),
      ## getRatioActive(dall, "Puducherry", lag = LAG, smooth = LAMBDA),
      getRatioActive(dall, "Tamil Nadu", lag = LAG, smooth = LAMBDA),
      getRatioActive(dall, "West Bengal", lag = LAG, smooth = LAMBDA)) |>
         xyplot(x = rcases ~ date | where, subset = date > as.Date("2020-07-01"),
                ylab = "Ratio of last two 1-week increments in confirmed cases",
                ## ylim = extendrange(c(0, 3)),
                prepanel = prepanelq,
                layout = c(NA, 2),
                abline = list(h = 1, lwd = 2, col = "grey"),
                type = "l", pch = 16, grid = TRUE)
```

