---
layout: remark
title: Measuring How Fast COVID Infections are Spreading in India
author: Deepayan Sarkar
mathjax: true
---

# Background

* Trying to measure growth rate rather than absolute number of cases

* The usual motivations:

	- Want something that can be compared across time and location
	
	- Something that should not depend on degree of systematic under-reporting

	- Something that should not be overly affected by daily fluctuations and patterns


* Measure used: number of new cases in _last seven days_ divided by number in _previous seven days_ 

* Anything higher than 1 is bad on a sustained basis, but more than 2 is _really_ bad



```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, autodep = TRUE,
               comment = "", warning = TRUE, message = FALSE, prompt = FALSE,
               dev.args = list(pointsize = 16), echo = FALSE,
               fig.width = 15, fig.height = 8, dpi = 96, fig.path='figures/incratio-')
options(warnPartialMatchDollar = FALSE, width = 100)
library(lattice)
library(RColorBrewer)
library(latticeExtra)
bdark <- RColorBrewer::brewer.pal(n = 8, name = "Dark2")
ct <- custom.theme(symbol = bdark, fill = adjustcolor(bdark, alpha.f = 0.5))
ct$strip.background$col <- "grey90"
ct$strip.border$col <- "grey50"
ct$axis.line$col <- "grey50"
lattice.options(default.theme = ct, default.args = list(as.table = TRUE))
```




```{r}
## source("tregularize.R")
source("covid-utils.R")
TARGET <- "districts.csv"
## To download the latest version, delete the file and run again
if (!file.exists(TARGET))
    download.file("https://api.covid19india.org/csv/latest/districts.csv",
                  destfile = TARGET)
dall <- read.csv(TARGET)
dall <-
    within(dall,
    {
        Date  <- as.Date(Date)
    })
```

.center[

[ Updated: `r {max(dall$Date)}` ]

]

---

# Cities: Bengaluru, Chennai, Delhi, Kolkata, Mumbai


```{r}
LAMBDA <- 0
LAG <- 7
rbind(getRatio(dall, district = "Delhi", lag = LAG),
      getRatio(dall, district = "Mumbai", lag = LAG),
      getRatio(dall, district = "Chennai", lag = LAG),
      getRatio(dall, district = "Bengaluru Urban", lag = LAG),
      getRatio(dall, district = "Kolkata", lag = LAG)) |>
         xyplot(x = rcases ~ date | where, subset = date > as.Date("2020-07-01"),
                ylab = "Ratio of last two 1-week increments in confirmed cases",
                ylim = extendrange(c(0, 3)), ## prepanel = prepanelq,
                abline = list(h = 1, lwd = 2, col = "grey"),
                type = "l", pch = 16, grid = TRUE)
```

---

# States: Delhi, Karnataka, Maharashtra, West Bengal

```{r}
LAMBDA <- FALSE
LAG <- 7
rbind(getRatio(dall, "Delhi", lag = LAG, smooth = LAMBDA),
      getRatio(dall, "Maharashtra", lag = LAG, smooth = LAMBDA),
      getRatio(dall, "Karnataka", lag = LAG, smooth = LAMBDA),
      getRatio(dall, "West Bengal", lag = LAG, smooth = LAMBDA)) |>
         xyplot(x = rcases ~ date | where, subset = date > as.Date("2020-07-01"),
                ylab = "Ratio of last two 1-week increments in confirmed cases",
                ## ylim = extendrange(c(0, 3)),
                prepanel = prepanelq,
                layout = c(NA, 2),
                abline = list(h = 1, lwd = 2, col = "grey"),
                type = "l", pch = 16, grid = TRUE)
```


---

# Current growth rates by state

.scrollable500[

```{r}
LAMBDA <- FALSE
LAG <- 7
latestRatio <- function(s, min.cases = 100)
{
    r <- getRatio(dall, state = s, lag = LAG, smooth = LAMBDA)
    c(round(rev(r$rcases)[c(1, 8)], 2),
      rev(r$icases)[c(1, 8)])
}
states <- unique(dall$State)
rstates <- as.data.frame(t(sapply(states, latestRatio)))
names(rstates) <-
    c("Current Rate", "Rate One week ago",
      "Cases in last 7 days", "Cases in previous 7 days")
kable(rstates[ order(rstates[[3]], decreasing = TRUE), ])
```

]


```{r}
states <- rownames(rstates)[is.finite(rstates[[1]])]
LAMBDA <- FALSE
LAG <- 7
slist <- lapply(sort(states),
                function(s) getRatio(dall, s, lag = LAG, smooth = LAMBDA, death.lag = 7))
sdf <- do.call(rbind, slist)
p <- xyplot(x = rcases ~ date | where, data = sdf,
            subset = (total >= 1e5 & date > as.Date("2020-07-01") & 
                      !(where %in% c("Delhi", "Maharashtra", "Karnataka", "West Bengal"))),
            ylab = "Ratio of last two 1-week increments in confirmed cases",
            ylim = extendrange(c(0, 3)),
            prepanel = prepanelq,
            abline = list(h = 1, lwd = 2, col = "grey"),
            type = "l", pch = 16, grid = TRUE)
N <- dim(p)
```

---

layout: true

# All other states with at least 100000 confirmed cases

---

```{r}
if (N > 6) p[1:6] else p
```

---

```{r}
if (N > 6) {
    if (N > 12) p[7:12] else p[7:N]
}
```

---

```{r}
if (N > 12) {
    if (N > 18) p[13:18] else p[13:N]
}
```

```{r}
if (N > 18) {
    p[19:N]
}
```

---

layout: false

# All election states have seen a surge

```{r}
LAMBDA <- FALSE
LAG <- 7
rbind(getRatio(dall, "Assam", lag = LAG, smooth = LAMBDA),
      getRatio(dall, "Kerala", lag = LAG, smooth = LAMBDA),
      ## getRatio(dall, "Puducherry", lag = LAG, smooth = LAMBDA),
      getRatio(dall, "Tamil Nadu", lag = LAG, smooth = LAMBDA),
      getRatio(dall, "West Bengal", lag = LAG, smooth = LAMBDA)) |>
         xyplot(x = rcases ~ date | where, subset = date > as.Date("2020-07-01"),
                ylab = "Ratio of last two 1-week increments in confirmed cases",
                ylim = extendrange(c(0, 3)),
                prepanel = prepanelq,
                layout = c(NA, 2),
                abline = list(h = 1, lwd = 2, col = "grey"),
                type = "l", pch = 16, grid = TRUE)
```

---

# Urban vs Rural

- In many states, COVID numbers were dominated by urban areas

- Has the pattern of growth been similar in urban vs rural?

- Has the pattern changed, especially due to the elections in March and April?

- In the following plots, "urban" districts are those within a state
  that cumulatively accounted for more than 50% of cases till March 1

---

# West Bengal


```{r}
LAMBDA <- FALSE
LAG <- 7
getRatioUR(dall, "West Bengal", lag = LAG, smooth = LAMBDA, cutoff.pc = 50) |>
    xyplot(x = rurban + rrural ~ date, # main = "West Bengal",
           subset = date > as.Date("2020-07-01"),
           main = "Urban = Kolkata / North 24 Parganas / South 24 Parganas",
           auto.key = list(lines = TRUE, points = FALSE, columns = 2,
                           text = c("Urban", "Rural")),
           ylab = "Ratio of last two 1-week increments in confirmed cases",
           ylim = extendrange(c(0, 3)),
           prepanel = prepanelq,
           panel = function(...) {
               panel.abline(v = as.Date(c("2020-08-01", "2021-03-01")), 
                            col = "grey50", lwd = 3)
               panel.xyplot(...)
               panel.text(x = as.Date(c("2020-08-01", "2021-03-01")) + 2,
                          y = 2,
                          labels = c("Unlock 3.0",
                                     "Election dates announced"),
                          col = "grey50", lwd = 3, srt = 90, pos = 1)
           },
           abline = list(h = 1, lwd = 2, col = "grey"),
           type = "o", pch = ".", cex = 2, grid = TRUE)
```

---

# Tamil Nadu 



```{r}
LAMBDA <- FALSE
LAG <- 7
getRatioUR(dall, "Tamil Nadu", lag = LAG, smooth = LAMBDA, cutoff.pc = 50) |>
    xyplot(x = rurban + rrural ~ date, # main = "Tamil Nadu",
           subset = date > as.Date("2020-07-01"),
           main = "Urban = Chennai / Coimbatore / Chengalpattu / Thiruvallur / Salem / Kancheepuram",
           auto.key = list(lines = TRUE, points = FALSE, columns = 2,
                           text = c("Urban", "Rural")),
           ylab = "Ratio of last two 1-week increments in confirmed cases",
           ylim = extendrange(c(0, 3)),
           prepanel = prepanelq,
           panel = function(...) {
               panel.abline(v = as.Date(c("2020-08-01", "2021-03-01")), 
                            col = "grey50", lwd = 3)
               panel.xyplot(...)
               panel.text(x = as.Date(c("2020-08-01", "2021-03-01")) + 2,
                          y = 2,
                          labels = c("Unlock 3.0",
                                     "Election dates announced"),
                          col = "grey50", lwd = 3, srt = 90, pos = 1)
           },
           abline = list(h = 1, lwd = 2, col = "grey"),
           type = "o", pch = ".", cex = 2, grid = TRUE)
```


---

# Kerala



```{r}
LAMBDA <- FALSE
LAG <- 7
getRatioUR(dall, "Kerala", lag = LAG, smooth = LAMBDA, cutoff.pc = 50) |>
    xyplot(x = rurban + rrural ~ date, # main = "Kerala",
           subset = date > as.Date("2020-07-01"),
           main = "Urban = Ernakulam / Kozhikode / Malappuram / Thiruvananthapuram / Thrissur",
           auto.key = list(lines = TRUE, points = FALSE, columns = 2,
                           text = c("Urban", "Rural")),
           ylab = "Ratio of last two 1-week increments in confirmed cases",
           ylim = extendrange(c(0, 3)),
           prepanel = prepanelq,
           panel = function(...) {
               panel.abline(v = as.Date(c("2020-08-01", "2021-03-01")), 
                            col = "grey50", lwd = 3)
               panel.xyplot(...)
               panel.text(x = as.Date(c("2020-08-01", "2021-03-01")) + 2,
                          y = 2,
                          labels = c("Unlock 3.0",
                                     "Election dates announced"),
                          col = "grey50", lwd = 3, srt = 90, pos = 1)
           },
           abline = list(h = 1, lwd = 2, col = "grey"),
           type = "o", pch = ".", cex = 2, grid = TRUE)
```

---

# Maharashtra



```{r}
LAMBDA <- FALSE
LAG <- 7
getRatioUR(dall, "Maharashtra", lag = LAG, smooth = LAMBDA, cutoff.pc = 50) |>
    xyplot(x = rurban + rrural ~ date, # main = "Maharashtra",
           subset = date > as.Date("2020-07-01"),
           main = "Urban = Pune / Mumbai / Thane / Nagpur",
           auto.key = list(lines = TRUE, points = FALSE, columns = 2,
                           text = c("Urban", "Rural")),
           ylab = "Ratio of last two 1-week increments in confirmed cases",
           ylim = extendrange(c(0, 3)),
           prepanel = prepanelq,
           abline = list(h = 1, lwd = 2, col = "grey"),
           type = "o", pch = ".", cex = 2, grid = TRUE)
```

---

# Karnataka



```{r}
LAMBDA <- FALSE
LAG <- 7
getRatioUR(dall, "Karnataka", lag = LAG, smooth = LAMBDA, cutoff.pc = 50) |>
    xyplot(x = rurban + rrural ~ date, # main = "Karnataka",
           subset = date > as.Date("2020-07-01"),
           main = "Urban = Bengaluru Urban / Mysuru / Ballari",
           auto.key = list(lines = TRUE, points = FALSE, columns = 2,
                           text = c("Urban", "Rural")),
           ylab = "Ratio of last two 1-week increments in confirmed cases",
           ylim = extendrange(c(0, 3)),
           prepanel = prepanelq,
           abline = list(h = 1, lwd = 2, col = "grey"),
           type = "o", pch = ".", cex = 2, grid = TRUE)
```

---

# Data quality: estimating death rate

- Kerala and Karnataka seem not to have controlled their growth

- In contrast, most North Indian states seem to be "improving" despite no obvious explanation

- Could this be due to under-reporting, whether due to strain on
  testing infrastructure or other reasons?

- One way to indirectly check this is to look at deaths vs cases

- The premise is that deaths will be reported more reliably than cases

- To account for lag, we compare new deaths in a 7-day window with new
  cases in the previous 7-day window

- Unfortunately, this only shows that data quality has been inconsistent
  throughout in most states

---

layout: true

# Death rate (with 7-day lag) 

---


```{r}
pd <- xyplot(x = I(ideaths / icases.lagged) ~ date | where, data = sdf,
            subset = total >= 1e5 & date > as.Date("2020-08-01"),
            ylab = "Ratio of new deaths (over 7 days) and new cases (over previous 7 days)",
            ylim = extendrange(c(0, 0.05)),
            prepanel = prepanelq,
            ## abline = list(h = 1, lwd = 2, col = "grey"),
            type = "l", pch = 16, grid = TRUE)
N <- dim(pd)
if (N > 24) message("N = ", N, " > 21, adjust next plot")
```


```{r}
update(pd[1:12], layout = c(3, 4))
```

---


```{r}
update(pd[13:N], layout = c(3, 4))
```


