---
layout: remark
title: Measuring How Fast COVID Infections are Spreading in India
author: Deepayan Sarkar
mathjax: true
---

class: center middle

# Measuring How Fast COVID Infections are Spreading in India


```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE, prompt = FALSE,
               dev.args = list(pointsize = 16), echo = FALSE,
               fig.width = 15, fig.height = 9, dpi = 96, fig.path='figures/incratio-')
options(warnPartialMatchDollar = FALSE, width = 100)
library(lattice)
library(RColorBrewer)
library(latticeExtra)
bpaired <- RColorBrewer::brewer.pal(n = 12, name = "Paired")
ct <- custom.theme(symbol = bpaired[c(FALSE, TRUE)], fill = bpaired[c(TRUE, FALSE)])
ct$strip.background$col <- "grey90"
ct$strip.border$col <- "grey50"
ct$axis.line$col <- "grey50"
lattice.options(default.theme = ct, default.args = list(as.table = TRUE))
```

---

# Background

* Trying to measure growth rate rather than absolute number of cases

* The usual motivations:

	- Want something that can be compared across time and location
	
	- Something that should not depend on degree of systematic under-reporting

	- Somthing that should not be affected by daily fluctuations and patterns


* Measure used: number of new cases in _last seven days_ divided by number in _previous seven days_ 


```{r}
source("tregularize.R")
source("covid-utils.R")
TARGET <- "districts.csv"
## To download the latest version, delete the file and run again
if (!file.exists(TARGET))
    download.file("https://api.covid19india.org/csv/latest/districts.csv",
                  destfile = TARGET)
dall <- read.csv(TARGET)
dall <-
    within(dall,
    {
        Date  <- as.Date(Date)
    })

```

---

# Delhi, Mumbai, Kolkata, Bengaluru


```{r}
LAMBDA <- 0
LAG <- 7
rbind(getRatio(dall, district = "Delhi", lag = LAG),
      getRatio(dall, district = "Mumbai", lag = LAG),
      getRatio(dall, district = "Chennai", lag = LAG),
      getRatio(dall, district = "Bengaluru Urban", lag = LAG),
      getRatio(dall, district = "Kolkata", lag = LAG)) |>
         xyplot(x = rcases ~ date | where, 
                ylab = "Ratio of last two 1-week increments in confirmed cases",
                ylim = extendrange(c(0, 3)), ## prepanel = prepanelq,
                abline = list(h = 1, lwd = 2, col = "grey"),
                type = "l", pch = 16, grid = TRUE)
```

---

## Delhi, Maharashtra, Karnataka, West Bengal

```{r}
LAMBDA <- FALSE
LAG <- 7
rbind(getRatio(dall, "Delhi", lag = LAG, smooth = LAMBDA),
      getRatio(dall, "Maharashtra", lag = LAG, smooth = LAMBDA),
      getRatio(dall, "Karnataka", lag = LAG, smooth = LAMBDA),
      getRatio(dall, "West Bengal", lag = LAG, smooth = LAMBDA)) |>
         xyplot(x = rcases ~ date | where, 
                ylab = "Ratio of last two 1-week increments in confirmed cases",
                ## ylim = extendrange(c(0, 3)),
                prepanel = prepanelq,
                abline = list(h = 1, lwd = 2, col = "grey"),
                type = "l", pch = 16, grid = TRUE)
```


---

# Current growth rates by state

.scrollable500[

```{r}
LAMBDA <- FALSE
LAG <- 7
latestRatio <- function(s, min.cases = 100)
{
    r <- getRatio(dall, state = s, lag = LAG, smooth = LAMBDA)
    round(rev(r$rcases)[c(1, 8)], 2)
}
states <- unique(dall$State)
rstates <- as.data.frame(t(sapply(states, latestRatio)))
names(rstates) <- c("current", "last_week")
kable(rstates[ order(rstates$current, decreasing = TRUE), ])
```

]


```{r}
states <- rownames(rstates)[is.finite(rstates$current)]
states <- setdiff(states, c("Delhi", "Maharashtra", "Karnataka", "West Bengal"))
LAMBDA <- FALSE
LAG <- 7
slist <- lapply(sort(states),
                function(s) getRatio(dall, s, lag = LAG, smooth = LAMBDA))
sdf <- do.call(rbind, slist)
p <- xyplot(x = rcases ~ date | where, data = sdf, subset = total >= 1e5,
            ylab = "Ratio of last two 1-week increments in confirmed cases",
            ylim = extendrange(c(0, 3)),
            prepanel = prepanelq,
            abline = list(h = 1, lwd = 2, col = "grey"),
            type = "l", pch = 16, grid = TRUE)
N <- dim(p)
```

---

layout: true

# All other states with at least 100000 confirmed cases

---

```{r}
if (N > 6) p[1:6] else p
```

---

```{r}
if (N > 6) {
    if (N > 12) p[7:12] else p[7:N]
}
```

---

```{r}
if (N > 12) {
    if (N > 18) p[13:18] else p[13:N]
}
```

---

```{r}
if (N > 18) {
    p[19:N]
}
```

