---
layout: default
title: 'A Simulated Infection Model'
author: Deepayan Sarkar
---


```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
opts_chunk$set(cache = FALSE, cache.path='~/knitr-cache/covid19/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
               knitr.table.format = "html",
			   ## dev.args = list(pointsize = 12),
               fig.width = 12, fig.height = 12, dpi = 120, fig.path='figures/deaths-')
library(lattice)
library(RColorBrewer)
library(latticeExtra)
options(stringsAsFactors = FALSE)
## bpaired <- RColorBrewer::brewer.pal(n = 12, name = "Paired")[-1]
## ct <- custom.theme(symbol = bpaired[c(FALSE, TRUE)], fill = bpaired[c(TRUE, FALSE)])
## ct$strip.background$col <- "grey90"
## ct$strip.border$col <- "grey50"
## ct$axis.line$col <- "grey50"
## lattice.options(default.theme = ct)
```

This is a very simple infection model developed with Krishanu Maulik.

A population initially consists of $n$ individuals, who live on a $M
\times N$ grid. Each individual has some attributes that are fixed and
some that evolve over time in a Markovian way:

- Fixed attributes:

	- home: $(u, v)$ where $1 \leq u \leq M, 1 \leq v \leq N$. Home
      coordinates (starting point)

	- type: constrained | mobile

	- radius: Integer. How far a constrained individual can move from
      home (Manhattan distance). Infinity for mobile individuals.
	
	- will_die: TRUE | FALSE. whether the person will die if infected (otherwise cured)
	
	- iduration: Integer. Number of days the individual
      remains contagious since infection.


- Time-varying attributes:

	- location $(x, y)$ where $1 \leq x \leq M, 1 \leq y \leq N$

	- status: susceptible | infected | cured | dead

	- days_infected: number of days since infection

	- static: TRUE | FALSE (a static person doesn't move)


Generally a non-static person tries to move like simple symmetric
random walk (north / south / east / west with equal
probability). However, moves that violate location constraints are not
allowed (the individual remains where they are).

Optionally, individuals can switch between static and non-static
states, with some probability. A static person does not move. This
allows mobile persons to move a long way, but then wait in a specific
location for a while. (Not implemented yet.)

Some persons are initially infected. An infected person infects
another (susceptible) person if they are in the same location at the
same time (with some probability?)


## Setup

```{r}
M <- 5000
N <- 5000
n <- 50000
RADIUS <- 50
ninitial <- 50 # number initially infected
pconstrained <- 0.6
pdeath <- 0.02
mean_iduration <- 20 * 24 # 20 days
lsample <- function(n, p) sample(c(TRUE, FALSE), n, replace = TRUE, prob = c(p, 1-p))
state <-
    data.frame(u = sample(M, n, replace = TRUE),
               v = sample(N, n, replace = TRUE),
               constrained = lsample(n, pconstrained),
               will_die = lsample(n, pdeath),
               iduration = rpois(n, lambda = mean_iduration-1) + 1,
               static = FALSE,
               status = "susceptible",
               days_infected = 0)
```


## Initialize

```{r}
initial <- sample(n, ninitial)
state <- within(state,
{
    radius <- ifelse(constrained, RADIUS, Inf)
    status[initial] <- "infected"
    days_infected[initial] <- 1
    x <- u
    y <- v
})
```

## Evolution rule


```{r}
allowed_move <- function(x, y, u, v, radius)
{
    (abs(x - u) <= radius) & (abs(y - v) <= radius) & 
        (x >= 1) & (x <= M) & (y >= 1) & (y <= N)
}
evolve <- function(s)
{
    within(s,
    {
        dx <- sample(c(-1, 1), n, replace = TRUE)
        dy <- sample(c(-1, 1), n, replace = TRUE)
        allowed <- allowed_move(x + dx, y + dy, u, v, radius)
        x <- x + allowed * dx
        y <- y + allowed * dy
        rm(dx, dy, allowed)
        ## susceptible -> infected
        which.infected <- which(status == "infected")
        ## loop through infected individuals and mark others in same location
        for (i in which.infected)
        {
            status[ (x == x[i]) & (y == y[i]) & (status == "susceptible") ] <- "infected"
        }
        rm(i)
        ## infected -> dead / cured
        ## First, increase number days since infection
        days_infected[which.infected] <- days_infected[which.infected] + 1
        status[which.infected] <- # FIXME: check
            ifelse(days_infected[which.infected] >= iduration[which.infected],
                   ifelse(will_die[which.infected], "dead", "cured"), "infected")
        rm(which.infected)

    })
}
```

## Simulation


```{r}
for (i in 1:(100 * 24))
{
    state <- evolve(state)
    if (i %% 24 == 0)
    {
        print(xyplot(y ~ x, state, groups = status,
                     main = as.character(i / 24),
                     auto.key = list(columns = 2)))
        print(sum(state$status != "susceptible"))
    }
}
```

```{r}
cases <- c(72 , 86 , 104 , 116 , 137 , 161 , 188 , 208 , 233 , 265 ,
           292 , 322 , 360 , 392 , 441 , 485 , 527 , 566 , 619 , 679 ,
           722 , 768 , 821 , 888 , 938 , 993 , 1047 , 1119 , 1180 ,
           1236 , 1299 , 1366 , 1436 , 1505 , 1567 , 1634 , 1696 ,
           1775 , 1845 , 1908 , 1990 , 2068 , 2136 , 2203 , 2273 ,
           2349 , 2427 , 2489 , 2544 , 2601 , 2652)
plot(cases, type = "o", pch = 16)
```
