---
layout: remark
title: Expressions, Evaluation and Scoping in R
author: Deepayan Sarkar
mathjax: true
---


```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE, prompt = FALSE,
               dev.args = list(pointsize = 16),
               fig.width = 15, fig.height = 6, dpi = 96, fig.path='figures/eval-')
options(warnPartialMatchDollar = FALSE, width = 100)
set.seed(20191105)
```

---

## REPL

* R works using what is known as a REPL (Read-Eval-Print-Loop)

	* R waits for user input when it starts
	
	* User types input _expression_ and presses `Enter`
	
	* R reads input and tries to _evaluate_
	
	* Evaluation is either successful or produces an error
	
	* R starts to wait again for further user input
	
* Today's talk takes a closer look at _how the evaluation step works_

--

* Useful links:

	* [Source](eval-slides.rmd) of this talk (R + Markdown, similar to Rnw)
	
	* [Makefile](Makefile) containing conversion rules (using [knitr](https://cran.r-project.org/package=knitr) and [pandoc](https://pandoc.org/))

	* Extracted [R code](eval-slides.R)


---

layout: true

## Evaluation

---

* Consider the following expression that we wish to evaluate


```{r}
sqrt(x)
```

* This expression involves two "symbols", `sqrt` and `x`

* To be able to evaluate the expression, R must be able to "find" both symbols

```{r}
find("sqrt")
find("x") # not found
```


---

* So let us define `x` and try again

```{r}
x <- 10
sqrt(x)
x <- -1
sqrt(x)
```


---

```{r}
x <- -1+0i
sqrt(x)
x <- "-1"
sqrt(x)
```

---

* Let us check again where `sqrt` and `x` are found

```{r}
find("sqrt")
find("x")
```

---

* Suppose we want to define a "smarter" `sqrt()`

* First check what `sqrt()` does, so we can mimic:

```{r}
sqrt
```

* What does `.Primitive()` do?

* Check `help(.Primitive)` (we don't need to understand details)

---

* define a "smart" `sqrt()` as follows:

```{r}
sqrt <- function(x)
{
    orig.sqrt <- .Primitive("sqrt")
    if (is.character(x)) x <- as.numeric(x)
    if (is.numeric(x) && any(x < 0)) x <- complex(real = x, imaginary = 0)
    orig.sqrt(x)
}
```

* We now have

```{r}
x <- -1
sqrt(x)
```

---

layout: false

## Multiple symbol definitions

* Where is R finding these symbols?

```{r}
find("x")
find("sqrt")
```



* The "original" `sqrt()` is not overwritten; instead a new one is defined


---

## The search path

* Why is it using the `sqrt` in `".GlobalEnv"` and not the one in `"package:base"`?

```{r}
search()
```



* R searches for symbols sequentially in the "search path"


---

## Environments

* What exactly are the objects in the search path?

* They are known as "environments", which are central to how evaluation works in R

* In essence, an environment is a _collection of symbols_ (variables)
  bound to _certain values_
  
* Each environment usually also has an associated "enclosing
  environment" or "parent environment"

* `.GlobalEnv` is the environment where user-defined variables are stored 

---

## Environments

* The chain of enclosing environments starting from `.GlobalEnv` is
  identical to the search path


```{r}
e <- .GlobalEnv
while (TRUE) { str(e, give.attr = FALSE); e <- parent.env(e) }
search() # compare
```


---

## Creating and manipulating environments

* `new.env()` explicitly creates environments

* Variables inside environments can be accessed using `$`


```{r,fig.height = 5}
e1 <- new.env()
e1$x <- seq(0, 2 * pi, length.out = 101)
e1$y <- cos(e1$x)
plot(y ~ x, data = e1, type = "l", ylim = c(-1, 1))
```



---

## Environments are not duplicated when copied

* Environments are an exception to R's duplicate-on-copy rule

* A copy of an environment is the same environment, not a new one

```{r}
e2 <- e1
range(e1$y)
e2$y <- abs(e2$y)
range(e1$y) # changes even though e1$y has not been directly modified 
```


---

## Inspecting the variables in an environment

* The names of objects available in an environment can be obtained by `ls()`

```{r}
ls(e2)
ls.str(e2) ## summary of all objects in e2
```


---

## Environments and functions

* Environments are central to how R functions work

* Functions in R are actually ["closures"](https://en.wikipedia.org/wiki/Closure_(computer_programming)

* In addition to the argument list and body (code), a function also
  has an associated environment

* This is usually the environment in which the function was defined

  
```{r}
environment(sqrt)
environment(base::sqrt) # special "primitive" (C) function, so no environment
environment(hist)
```


---

## What happens when a function is executed?

* Every time a function is invoked, a new environment is created

* This is the "evaluation environment" and contains the local variables of the function

* The environment enclosing this "evaluation environment" is the
  environment of the function
--

* While the function is running, evaluation tries to find symbols as follows:

	* First, in the evaluation environment
	
	* Next, in its enclosing environment, the environment of the function
	
	* Then, in the enclosing environment of that environment, and so on

* This behaviour is called "lexical scoping"
--

* Evaluation in the global environment follows the same rule (hence the search path)



---

## Why are these details important?

* Depending on scoping rules (for non-local symbols) is often risky


```{r}
boxcox <- function(x) # applies Box-Cox transformation on data 'x'
{
	if (lambda == 0) log(x)
	else (x^lambda - 1) / lambda
}
b <- function(x) # computes histogram bins given data 'x' 
{
    n <- sum(is.finite(x))
    r <- extendrange(x)
    seq(r[1], r[2], length.out = round(sqrt(n))+1)
}
```

```{r}
x <- rlnorm(1000)
hist(boxcox(x), breaks = b)
```


---

## Why are these details important?

* Works if we define `lambda`


```{r}
lambda <- 0
hist(boxcox(x), breaks = b)  # notice that 'b' is a function! (See ?hist)
```




---

## Why are these details important?

* Normally, depending on this kind of scoping behaviour should be unnecessary

* In fact, a well-designed function should not depend on arbitrary variables

* However, there is one situation where this evaluation rule becomes important

* When a function creates and returns another function!


---

## Do we need functions that create functions?

* Have you encountered such functions?
--

* Probably not very commonly observed...

* But only because we are limited in our thinking

* We have already used functions as arguments to other functions

* Thinking "functionally" opens up many other interesting approaches

* This is particularly useful for quick prototyping


---

## Do we need functions that create functions?

* A somewhat silly extension of the example above:

```{r}
fboxcox <- function(lambda)
{
	f <- function(x)
	{
		if (lambda == 0) log(x)
		else (x^lambda - 1) / lambda
	}
    return(f)
}
```

```{r}
boxcox.sqrt <- fboxcox(0.5)
boxcox.log <- fboxcox(0)
```



---

## Do we need functions that create functions?

```{r}
par(mfrow = c(1, 2))
hist(boxcox.sqrt(x), breaks = b)
hist(boxcox.log(x), breaks = b)
```




---

## Functions returned by functions

* Note that `boxcox.sqrt` and `boxcox.log` are both functions (created by
  another function `fboxcox`)

```{r}
boxcox.sqrt
boxcox.log
```


* ... that appear to be identical


---

## Functions returned by functions

* But they have different "environments"

```{r}
environment(boxcox.sqrt)
environment(boxcox.log)
```


---

## Functions returned by functions

* This is because they were created in different executions of `fboxcox()`

* Although not immediately apparent, these environments retain the value of `lambda`

```{r}
ls.str(environment(boxcox.sqrt))
ls.str(environment(boxcox.log))
```

---

## Functions returned by functions

* Without lexical scoping, this would not have been possible

* Another example from [The R Paper](https://www.stat.auckland.ac.nz/~ihaka/downloads/R-paper.pdf):

```{r}
y <- 123
f <- function(x) {
    y <- x * x
	g <- function() print(y)
	g()
}
f(10)
```



* S, which is similar to R but does not have lexical scoping, would print `123`



---

## More serious examples

* Functions are actually quite commonly encountered in probability and statistics

* Examples: density function, CDF, likelihood function

---

## The empirical CDF

* Consider the empirical cumulative distribution function (ECDF):

```{r}
ECDF <- function(x)
{
	x <- x[is.finite(x)]
	function(u) {
	    sum(x <= u) / length(x)
	}
}
x <- rlnorm(50)
F.sqrt <- ECDF(boxcox.sqrt(x))
F.log <- ECDF(boxcox.log(x))
```

* Evalute ECDF at certain values

```{r}
sapply(-5:5, F.sqrt)
sapply(-5:5, F.log)
```

---

## The empirical CDF

* Many R tools are designed to work with vectorized functions

* The `Vectorize()` function creates a vectorized function from a non-vectorized one


```{r}
FF.log <- Vectorize(F.log)
plot(FF.log, from = -4, to = 4, n = 1000)
```



---

## The empirical CDF

* But where is the data `x`?

```{r}
F.log
```



* It is still available in the environment that _created_ `F.log`

```{r}
environment(F.log)$x
```

* In fact, R has has a built-in `ecdf()` function that works similarly



---

## Objective functions (to be optimized)

* Another common use case: optimization of a loss function

* Consider the following general approach to linear regression given data $(x_i, y_i), i = 1,\dotsc,n$

<div>
$$\hat{\beta} = \arg \min_{\beta} \sum_{i=1}^n L(y_i - \beta_0 - \beta_1 x_i) = \arg \min_{\beta} f(\beta)$$
</div>

* Common choices for $L$ are

	* $L(u) = u^2$

	* $L(u) = | u |$

	* <div>$L(u) = \begin{cases} 
	             u^2 & \lvert u \rvert \leq c \\ 
	             c (2 | u | - c) & \text{otherwise} 
			  \end{cases}$</div>


---

## Objective functions (to be optimized)

* The following functions takes data and $L$, and creates a suitable objective function

```{r}
make.objective <- function(x, y, L = abs)
{
	function(beta) {
	    sum(L(y - beta[1] - beta[2] * x))
	}
}
data(phones, package = "MASS")
obj.lad <- with(phones, make.objective(year, calls, L = abs))
obj.lse <- with(phones, make.objective(year, calls, L = function(u) u*u))
```


---

## Objective functions (to be optimized)


```{r}
obj.lad
environment(obj.lse)$L
environment(obj.lad)$L
```

---

## Objective functions (to be optimized)


```{r}
fm.lad <- optim(obj.lad, par = c(0,0))
fm.lse <- optim(obj.lse, par = c(0,0))
str(fm.lad)
str(fm.lse)
```


---

## Objective functions (to be optimized)


```{r,fig.height=7}
plot(calls ~ year, phones)
abline(fm.lad$par, col = "blue")
abline(fm.lse$par, col = "red")
```




---

## A more complex example

* Suppose we want to plot a histogram

* One important parameter is the bin width $h$

* Fixing starting point, what is an optimal choice?

* One option: [cross validation](http://onlinestatbook.com/stat_sim/histogram/index.html)


---

## Cross validation: summary


$\newcommand{\fhath}{\hat{f}_h}$

* Denote the histogram for bin width $h$ by $\fhath$, and the unknown density by $f$

* Want to choose $h$ to minimize integrated square error

$$
\int \left[ \fhath(x) - f(x) \right]^2 dx = 
\int \fhath(x)^2 dx - 2 \int \fhath(x)\, f(x)\, dx + \int f(x)^2 dx
$$

* The third term is unknown but constant (as $h$ varies), so can be dropped

* The first term can be calculated from $\fhath$ (which is piecewise constant)

* The second term involves the unknown $f$, but the integral can be viewed as $E(\fhath(X))$ when $X \sim f$


---

## Cross validation: summary

* Leave-one-out estimator: average of ${\fhath}_{(-i)}(x_i)$

* The estimated integrated square error (without the constant term) simplifies to

$$
\frac{2}{(n-1) h} - \frac{n+1}{(n-1) n^2 h} \sum_{k} c_k^2
$$

* where $c_k$ are bin counts


---

## Cross validation: implementation


```{r}
chooseBins <- function(x, p = 0.25)
{
    x <- x[is.finite(x)]
    n <- length(x)
    r <- extendrange(x)
    d <- r[2] - r[1]
    CVE <- function(B)
    {
        h <- d / B
        breaks <- seq(r[1], r[2], length.out = B+1)
        k <- findInterval(x, breaks) # runtime O(n * log(B))
        counts <- table(factor(k, levels = 1:B))
        2 / ((n-1)*h) - sum(counts^2) * (n+1) / ((n-1) * n^2 * h)
    }
    Bvec <- seq(floor(sqrt(n)*p), ceiling(sqrt(n)/p))
    list(range = r, B = Bvec, CVE = unname(sapply(Bvec, CVE)))
}
```


---

## Cross validation: implementation


```{r}
xx <- c(rnorm(500), rnorm(300, mean = 6, sd = 1.5))
ll <- chooseBins(xx, p = 0.125)
str(ll)
```


---

## Cross validation: implementation


```{r}
with(ll, {
    plot(CVE ~ B, type = "o")
    abline(v = B[which.min(CVE)])
})
```



---

## Cross validation: comparison with other "rules"


```{r}
(nclass.cv <- with(ll, B[which.min(CVE)]))
nclass.Sturges(xx)
nclass.FD(xx)
nclass.scott(xx)
```


---

## Cross validation: Resulting histogram


```{r,fig.height=5.5}
breaks <- seq(ll$range[1], ll$range[2], length.out = nclass.cv + 1)
hist(xx, breaks = breaks)
```

* For more interesting examples, see Gentleman and Ihaka,
  [Lexical Scope and Statistical Computing](https://www.stat.auckland.ac.nz/~ihaka/downloads/lexical.pdf)




---

## Related topic: non-standard evaluation


* Again, consider an expression that will cause an error when evaluated


```{r}
rm(list = c("x", "y")) # remove x, y if defined earlier
plot(x, y, type = "l")
```



---

## Quoted (unevaluated) expressions

* R can actually work with an expression without evaluating it


```{r}
e <- quote(plot(x, y, type = "l"))
e
e[[1]]
e[[2]]
e[[3]]
```

---

## Evaluating Quoted expressions

* A quoted expression can be evaluated using `eval()`


```{r}
eval(e)
```

* Fails because current environment (`.GlobalEnv`) does not contain variables `x` and `y`

* But Recall: Environment `e2` defined earlier contains 

```{r}
ls.str(e2)
```


---

## Evaluating Quoted expressions

```{r}
eval(e, envir = e2)
```



---

## Can we re-implement the `replicate()` function?

* This is how functions like `replicate()` and `with()` work

```{r}
my.replicate <- function(N, e)
{
    ans <- numeric(N)
    for (i in 1:N) ans[[i]] <- eval(e)
    ans
}
```


* But we have to be careful to use it

```{r}
my.replicate(5, max(rnorm(10))) # not what we wanted
my.replicate(5, quote(max(rnorm(10))))
```

---

## Non-standard evaluation

* The `substitute()` trick: non-standard evaluation (using lazy argument evaluation)

```{r}
my.replicate <- function(N, e)
{
    expr <- substitute(e) # avoids the need to quote()
    ans <- numeric(N)
    for (i in 1:N) ans[[i]] <- eval(expr)
    ans
}
str(x <- my.replicate(1000, max(rnorm(20))))
```

---

## Lazy evaluation

* Lazy evaluation: function arguments are only evaluated when needed


```{r}
set.seed(20191105)
rm(sqrt) # remove the sqrt() we defined earlier
chooseOne <- function(a, b, u = runif(1))
{
    if (u < 0.5) a else b
}
chooseOne(sqrt(2), sqrt(-2)) # no warning message
chooseOne(sqrt(2), sqrt(-2)) # warning message
```


---

## Further reading

```{r, eval=FALSE}
? environment
? eval
? expression
? quote
? bquote
? substitute
? with
```

```{r}
with.default
replicate
```

