---
layout: remark
title: An Overview of R - 1
subtitle: Introductory Computer Programming
author: Deepayan Sarkar
mathjax: true
---


```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
opts_chunk$set(cache = TRUE, cache.path='~/knitr-cache/icp-roverview-1/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
               knitr.table.format = "html",
               fig.width = 15, fig.height = 6,
               dpi = 96, fig.path='figures/roverview-1-')
library(lattice)
lattice.options(default.args = list(as.table = TRUE), 
                default.theme = list(grid.pars = list(cex = 1.5)))
options(warnPartialMatchDollar = FALSE, width = 120)
```

# Software for Statistics

- Computation is an essential part of modern statistics

    - Handling large datasets 

    - Visualization

    - Simulation

    - Iterative methods 

- Many softwares are available for such computation, but we will focus on R
--

- Why R?

    - Available as [Free](https://en.wikipedia.org/wiki/Free_software_movement) / 
    	  [Open Source](https://en.wikipedia.org/wiki/The_Open_Source_Definition) Software

    - Very popular (both academia and industry)

    - Easy to try out on your own
--

- Also because I know R better than other languages

- Python and Julia are other good alternatives


---

# Outline

- Installing R (hopefully most of you have already done this)

- Basics of using R

- Some examples


---

# Installing R

* R is most commonly used as a [REPL](https://en.wikipedia.org/wiki/Read-eval-print_loop) (Read-Eval-Print-Loop)

	* When it is started, R Waits for user input

	* Evaluates and prints result

	* Waits for more input
--

* There are several different _interfaces_ to do this

* R itself works on many platforms (Windows, Mac, UNIX, Linux)

* Some interfaces are platform-specific, some work on most
--

* R and the interface may need to be installed separately


---

# Installing R

* Go to <https://cran.r-project.org/> (or choose a [mirror](https://cran.r-project.org/mirrors.html) first)

* Follow instructions depending on your platform (Windows for most of you)
--

* This will install R, as well as a default graphical interface on Windows and Mac
--

* I will recommend a different interface called
  [R Studio](https://www.rstudio.com/) that needs to be installed separately
  
* I personally use yet another interface called
  [ESS](https://ess.r-project.org/) with a general-purpose
  editor called [Emacs](https://www.gnu.org/software/emacs/)


---

# Running R

* Once installed, you can start the appropriate interface (or R directly) to get something like this:

```
R version 4.0.3 (2020-10-10) -- "Bunny-Wunnies Freak Out"
Copyright (C) 2020 The R Foundation for Statistical Computing
Platform: x86_64-apple-darwin19.6.0 (64-bit)

R is free software and comes with ABSOLUTELY NO WARRANTY.
You are welcome to redistribute it under certain conditions.
Type 'license()' or 'licence()' for distribution details.

R is a collaborative project with many contributors.
Type 'contributors()' for more information and
'citation()' on how to cite R or R packages in publications.

Type 'demo()' for some demos, 'help()' for on-line help, or
'help.start()' for an HTML browser interface to help.
Type 'q()' to quit R.

> 
```

* The `>` represents a _prompt_ indicating that R is waiting for input.

* The difficult part is to learn what to do next


---

# The R REPL essentially works like a calculator

```{r}
34 * 23
27 / 7
exp(2)
2^10
```

---

# R has standard mathematical functions

```{r}
sqrt(5 * 125)
log(120)
factorial(10)
log(factorial(10))
```

---

# R has standard mathematical functions

```{r}
choose(15, 5)
factorial(15) / (factorial(10) * factorial(5))
```
--

```{r}
choose(1500, 2)
factorial(1500) / (factorial(1498) * factorial(2))
```



---

# R supports variables

```{r}
x <- 2
y <- 10
x^y
y^x
factorial(y)
log(factorial(y), base = x)
```

---

# R can compute on vectors

```{r}
N <- 15
x <- seq(0, N)
N
x
choose(N, x)
```


---

# R has functions for statistical calculations

```{r}
p <- 0.25
choose(N, x) * p^x * (1-p)^(N-x)
dbinom(x, size = N, prob = p)
```

---

# R has functions that work on vectors

```{r}
p.x <- dbinom(x, size = N, prob = p)
sum(x * p.x) / sum(p.x)
N * p
```


---

# R can draw graphs

```{r}
plot(x, p.x, ylab = "Probability", pch = 16)
title(main = sprintf("Binomial(%g, %g)", N, p))
abline(h = 0, col = "grey")
```
\



---

# R can simulate random variables

```{r}
cards <- as.vector(outer(c("H", "D", "C", "S"), 1:13, paste))
cards
```
--

```{r}
sample(cards, 13)
sample(cards, 13)
```


---

# R can simulate random variables

```{r}
z <- rnorm(50, mean = 0, sd = 1)
z
```



---

# Example: random walk

```{r}
plot(1:100, type = "n", ylim = c(-25, 25), ylab = "")
for (i in 1:20) {
    z <- rnorm(100, mean = 0, sd = 1)
    lines(cumsum(z), col = sample(colors(), 1))
}
```
\



---

# R is in fact a full programming language

* Variables

* Functions

* Control flow structures

	* For loops, while loops

	* If-then-else (branching)
--

* Distinguishing features

	* Focus on _vectors_ and _vectorized operations_ 

	* Treatment of _functions_ at par with other object types


* We will see a few examples to illustrate what I mean by this in the next class

* First, a demo of using R using an [R script]() and an [R Markdown]() file 

