---
layout: remark
title: Basic usage of R - 1
subtitle: Introductory Computer Programming
author: Deepayan Sarkar
mathjax: true
---



# Basics of using R

R is more flexible than a regular calculator

* In fact, R is a full programming language

* Most standard data analysis methods are already implemented

* Can be extended by writing add-on packages

* Thousands of add-on packages are available



```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
opts_chunk$set(cache = TRUE, cache.path='~/knitr-cache/icp-rbasics-1/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
               knitr.table.format = "html",
               fig.width = 15, fig.height = 6, dpi = 96, fig.path='figures/rbasics-1-')
options(warnPartialMatchDollar = FALSE, width = 100)
```





---

# Major concepts

* Variables (in the context of programming)

* Data structures needed for data analyis

* Functions (set of instructions for performing a procedure)
---

# Variables

* Variables are symbols that may be associated with different values

* Computations involving variables are done using their current value

```{r}
x <- 10 # assignment
sqrt(x)
x <- -1
sqrt(x)
x <- -1+0i
sqrt(x)
```
---

# Data structures for data analysis


* Vectors

* Matrices 

* Data frames (a spreadsheet-like data set)

* Lists (general collection of objects)

---

# Atomic vectors

* Indexed collection of homogeneous scalars, can be 

    * Numeric / Integer

    * Categorical (factor)

    * Character

    * Logical (`TRUE` / `FALSE`)
	
* Missing values are allowed, indicated as `NA`

* Elements are indexed starting from 1

* $i$th element of vector `x` can be extracted using `x[i]`

* There are also more sophisticated forms of (vector) indexing

---

# Atomic vectors: examples

```{r}
month.name # built-in 
x <- rnorm(10)
x
```
--


```{r}
str(x) # useful function
str(month.name)
```
---

# Atomic vectors: examples

```{r}
m <- sample(1:12, 30, rep = TRUE)
m
mf <- factor(m, levels = 1:12, labels = month.name)
mf
str(m)
str(mf)
```

---

# Atomic vectors

* "Scalars" are just vectors of length 1

```{r}
str(numeric(2))
str(numeric(1))
str(0)
```
---

# Atomic vectors


* Vectors can have length zero 

```{r}
numeric(0)
logical(0)
```




---

# Types of indexing

* Indexing refers to extracting subsets of vectors (or other kinds of data)

* R supports several kinds of indexing:

	* Indexing by a vector of positive integers

	* Indexing by a vector of negative integers
	
	* Indexing by a logical vector

	* Indexing by a vector of names

---

# Types of indexing

* The "standard" C-like indexing with a scalar (vector of length 1):

```{r}
month.name[2] # the first index is 1, not 0
```

* The "index" can also be an integer vector

```{r}
month.name[c(2, 4, 6, 9, 11)]
```

* Elements can be repeated

```{r}
month.name[c(2, 2, 6, 4, 6, 11)]
```


---

# Types of indexing

* "Out-of-bounds" indexing give `NA` (missing)

```{r}
month.name[13]
seq(1, by = 2, length.out = 8)
month.name[seq(1, by = 2, length.out = 8)]
```


---

# Types of indexing

* Negative integers omit the specified entries

```{r}
month.name[-2]
month.name[-c(2, 4, 6, 9, 11)]
```


* Cannot be mixed with positive integers

```{r}
month.name[c(2, -3)]
```
---

# Types of indexing

* Zero has a special meaning - doesn't select anything

```{r}
month.name[0]
month.name[integer(0)] ## same as empty index
month.name[c(1, 2, 0, 11, 12)]
month.name[-c(1, 2, 0, 11, 12)]
```
---

# Types of indexing

* Indexing by logical vector: select `TRUE` elements

```{r}
month.name[c(TRUE, FALSE, FALSE)] # index replicated
i <- substring(month.name, 1, 1) == "J"
i
month.name[i]
```

---

# Types of indexing

* Common use: extract subset satisfying a certain condition (also called "filtering")

```{r}
(x <- rnorm(20))
x > 0
x[x > 0]
mean(x[x > 0])
```


---

# Types of indexing

* Sometimes logical indexing can be replaced by integer indexing using `which()`

```{r}
i
which(i)
month.name[ which(i) ]
month.name[ -which(i) ] # same as month.name[ !i ]
```
---

# Types of indexing

* But be careful about zero-length indices

```{r}
which(substring(month.name, 1, 1) == "B")
month.name[ which( substring(month.name, 1, 1) == "B") ]
-which(substring(month.name, 1, 1) == "B") 
month.name[ -which( substring(month.name, 1, 1) == "B") ]
```

---

class: center middle

# Questions?



