---
layout: remark
title: Basic usage of R - 2
subtitle: Introductory Computer Programming
author: Deepayan Sarkar
mathjax: true
---

# Lists

* Lists are vectors with arbitrary types of components

* May or may not have names

* Usual vector indexing by `[` works in the usual way

* Individual elements can be extracted either by name (`$`) or by position `x[[i]]`


```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
opts_chunk$set(cache = TRUE, cache.path='~/knitr-cache/icp-rbasics-2/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
               knitr.table.format = "html",
               fig.width = 15, fig.height = 6, dpi = 96, fig.path='figures/rbasics-2-')
options(warnPartialMatchDollar = FALSE, width = 100)
```


---

# Lists: examples

```{r}
m <- sample(1:12, 30, rep = TRUE)
mf <- factor(m, levels = 1:12, labels = month.name)
ml <- list(imonth = m, fmonth = mf)
str(ml)
```
--


```{r}
ml$imonth
```
--


```{r}
ml[[2]]
```

<!-- ml[["fmonth"]] -->

---

# Lists: examples

* Lists are commonly used to return analysis results (more details later)

```{r}
tt <- t.test(rnorm(100))
str(tt, give.attr = FALSE)
tt$p.value
```


---

layout: true

# Common structures for statistical data

---

* Vectors, matrices / arrays: vectors with dimension

```{r}
VADeaths
dim(VADeaths)
```
---

* Indexing works in same way, but in two dimensions

```{r}
VADeaths[1:2, c(2, 3)]
```

* Example: Indexing by "empty" index (selects all) and name

```{r}
VADeaths[, c("Rural Male", "Rural Female")]
```

---

* Data frames: lists that also behave like a matrix

```{r}
str(mtcars)
dim(mtcars)
```

---

layout: true

# Data Frames

---

* Rectangular (matrix-like) structure

* Each column is (usually) a vector

* Different columns can have different types (unlike a matrix)

* Every column must have the same length

* Every column must have a name
--

* Most built-in data sets in R are data frames

---

* List-like behaviour: Columns can be extracted like a list
	
```{r}
mtcars$mpg
```

* Vector indexing extracts list elements

```{r}
head(mtcars[c(1, 4, 7)])
```

---

* Matrix-like behaviour: Two-dimensional indexing

```{r}
mtcars[1:6, c(1, 4, 7)]
mtcars[sample(nrow(mtcars), 6), c("mpg", "wt", "am", "gear")]
```

---

layout: true

# Data input / output

---

* Statistical data are usually structured like a spreadsheet (e.g., Excel)

* Typical approach: read data from spreadsheet file into data frame

* Easiest route:

    * R itself cannot read Excel files directly

    * Save as CSV file from Excel

    * Read with `read.csv()` or `read.table()` (more flexible)

* Alternative option:

    * Use "Import Dataset" menu item in R Studio (requires add-on packages)

---

* Data frames can be exported as a spreadsheet file using
  `write.csv()` or `write.table()`

```{r}
data(Cars93, package = "MASS")
write.csv(Cars93, file = "temp/cars93.csv") # export
```

* This creates a [text file](temp/cars93.csv) (that can be easily edited in Excel / LibreOffice)

.scrollable300[

```{bash,echo=FALSE}
cat cars93.csv
```

]

---

* We can use `read.csv()` to import it again

.scrollable400[

```{r}
cars <- read.csv("temp/cars93.csv") # import (path relative to working directory)
str(cars)
```

]

---

* These are the most basic data input / output functions

* There are many other other specialized functions

* Low-level utilties: `scan()`, `readLines()`, `readChar()`, `readBin()`

* Various packages provide import / export to formats used by other software

* R has its own "serialization" format using `save()` and `load()`


---

class: center middle

# Questions?

