---
layout: remark
title: Basic usage of R - 2
subtitle: Introductory Computer Programming
author: Deepayan Sarkar
mathjax: true
---

layout: true

# Functions

---

* Most useful things in R happen by calling functions

* Functions have one or more arguments

    * All arguments have names (available as variables inside the function)
--

	* Arguments may be compulsory or optional

	* Optional arguments usually have "default" values 
--

	* Arguments may or may not be named when the function is called
	
	* Unnamed arguments are matched by position

	* Optional arguments are usually named
	
* Functions normally also have a useful "return" value


```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
opts_chunk$set(cache = TRUE, cache.path='~/knitr-cache/icp-rbasics-3/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
               knitr.table.format = "html",
               fig.width = 15, fig.height = 6, dpi = 96, fig.path='figures/rbasics-3-')
options(warnPartialMatchDollar = FALSE, width = 100)
```



---

* For example, linear models are fit using the function `lm()`

```{r}
str(lm)
```

* These calls are equivalent

```{r}
fm1 <- lm(mpg ~ wt, mtcars, (am == 0))
fm2 <- lm(mpg ~ wt, data = mtcars, method = "qr", subset = (am == 0))
coef(fm1)
coef(fm2)
```

* Rule: named arguments are matched by name, remaining by position

---

* The return value of `lm()` is a list

.scrollable400[

```{r}
str(fm2)
```

]

---

* Individual elements can be extracted using list indexing

```{r}
fm2$coefficients
fm2$residuals
```

---

layout: true

# Attributes

---

* The names (derived from row names of the data) can be used as index 

```{r}
fm2$residuals["Chrysler Imperial"]
```

* The names associated with a vector can be obtained using `names()`

```{r}
names(fm2$residuals)
```

---

* These names are actually stored as an _attribute_ called "names"

```{r}
attr(fm2$residuals, "names")
```

* Attributes are arbitrary R objects that can be attached to any other object

* Typically used for programming convenience, normally not seen by users

* However, some attributes are "special"

---

* The "names" attribute can be extracted using the function `names()`

* `dimnames()` similarly gives row / column names for matrices and arrays

```{r}
dimnames(VADeaths)
attr(VADeaths, "dimnames")
```

---

* For example, column names can be obtained as

```{r}
dimnames(VADeaths)[[2]]
```

* There are convenient shortcuts called `rownames()` and `colnames()`

```{r}
colnames(VADeaths)
```

---

* In fact, we can easily verify that this is what `colnames()` is doing by printing it

```{r}
colnames
```

* All R functions can be easily inspected in this way

---

* Another very important attribute is "class"

* For example, the return value of `lm()` has class "lm"

```{r}
attr(fm2, "class")
```

* The class of an object can also be obtained using the function `class()`

```{r}
class(fm2)
```

---

* The class of an object can (usually) be "removed" by setting it to
  `NULL` 
  
* This is not something you should actually do!

```{r}
class(fm2) <- NULL
```

* Such objects will no longer have a "class" attribute

```{r}
attr(fm2, "class")
```

* However, it will still have a class (implicitly)

```{r}
class(fm2)
```

---

layout: true

# Class, generic functions, and methods

---

* The class of an object is fundamental to how R works

* Every object in R must have a class

* This is true even if the object does not have a class attribute

```{r}
class(colnames)
attr(colnames, "class")
class(VADeaths)
attr(VADeaths, "class")
```

---

* The main use of the class of an object is in how _generic functions_ behave

* Generic functions are intended to perform general tasks, like

	- `print()`

	- `plot()`

	- `summary()`

* But details of what these functions should do depends on the input


---

```{r}
print(VADeaths) 
print(fm1)
```

---

```{r}
summary(VADeaths) 
```

---

```{r}
summary(fm1)
```

---

* But recall that we removed the class attribute from `fm2`

```{r}
class(fm1)
class(fm2)
```

* Otherwise they represent the same model fit

* But the different class means that `print()` and `summary()` behave
  differently

---

```{r}
summary(fm2)
```

---

.scrollable400[

```{r}
print(fm2)
```

]

* This kind of customized output is achieved by _methods_

---

* Methods are specific implementations of a generic function customized to its input

* The appropriate method is chosen by looking at the _class_ of the input argument
--

* The methods available for a generic function can be obtained using the `methods()` function

```{r}
methods("summary")
```

```{r, eval=FALSE}
methods("print") # similar but much longer list
```

---

layout: false

# Getting help

* R has an extensive collection of functions (even excluding add-on packages)

* It is impossible for anyone to know them all, or remember details

* Fortunately, R also has an excellent help system
--

* Every function and dataset in R must be documented

* The documentation can be accessed by `help(name)` or `?name`

* For example: `help(seq)`, `help(summary)`, etc.

---

# Getting help

* How the help is shown depends on the _interface_ being used

* RStudio has a separate help tab which also allows searching
--

* However, before using the help system, you should know a bit about
  how methods are documented

---

# Help for generic functions and methods



---

# Other important features

* Getting help

* Help for generic functions and methods

* How matrices and arrays are implemented in R

<!-- 

* Homework --- read the following tutorials

	* [Introduction and Language Overview I](https://www.isid.ac.in/~deepayan/R-tutorials/labs/01_introduction_lab.pdf)

	* [Language Overview II](https://www.isid.ac.in/~deepayan/R-tutorials/labs/02_language_lab.pdf)

	* Supplementary files:
      [Gcsemv.txt](https://www.isid.ac.in/~deepayan/R-tutorials/data/Gcsemv.txt),
      [iris.xls](https://www.isid.ac.in/~deepayan/R-tutorials/data/iris.xls),

-->

class: center middle

# Questions?


