---
layout: remark
title: Basic usage of R - 4
subtitle: Introductory Computer Programming
author: Deepayan Sarkar
mathjax: true
---

layout: true

# Class, generic functions, and methods

---

* The class of an object is fundamental to how R works

* Every object in R must have a class

* This is true even if the object does not have a class attribute

```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
opts_chunk$set(cache = TRUE, cache.path='~/knitr-cache/icp-rbasics-4/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
               knitr.table.format = "html",
               fig.width = 15, fig.height = 6, dpi = 96, fig.path='figures/rbasics-4-')
options(warnPartialMatchDollar = FALSE, width = 100)
```


```{r}
class(colnames)
attr(colnames, "class")
class(VADeaths)
attr(VADeaths, "class")
```

---

* The main use of the class of an object is in how _generic functions_ behave

* Generic functions are intended to perform general tasks, like

	- `print()`

	- `plot()`

	- `summary()`

* But details of what these functions should do depends on the input


---

```{r}
print(VADeaths) 
fm1 <- lm(mpg ~ wt, mtcars, subset = (am == 0))
print(fm1)
```

---

```{r}
summary(VADeaths) 
```

---

```{r}
summary(fm1)
```

---

* But suppose we make a copy of `fm1` and remove the class attribute from it

```{r}
fm2 <- fm1
class(fm2) <- NULL
class(fm1)
class(fm2)
```

* `fm1` and `fm2` represent the same model fit

* But the different class means that `print()` and `summary()` behave
  differently

---

```{r}
summary(fm2)
```

---

.scrollable400[

```{r}
print(fm2)
```

]

* This kind of customized output is achieved by _methods_

---

* Methods are specific implementations of a generic function customized to its input

* The appropriate method is chosen by looking at the _class_ of the input argument
--

* The methods available for a generic function can be obtained using the `methods()` function

```{r}
methods("summary")
```

```{r, eval=FALSE}
methods("print") # similar but much longer list
```

---

* All available methods for a given class can be similarly obtained

```{r}
methods(class = "lm")
```

---

* The name of a specific method appears to have the form `generic.class`

* However, one should always call the generic function, not the method directly

* This is not OK:

```{r,eval=FALSE}
summary.lm(fm1)
```

* Instead, use

```{r,eval=FALSE}
summary(fm1)
```

* In fact, many methods cannot be called directly because they are "hidden"

---

layout: false

# Getting help

* R has an extensive collection of functions (even more if we include
  add-on packages)

* It is impossible for anyone to know them all, or remember details

* Fortunately, R also has an excellent help system
--

* Every function and dataset in R (and add-on packages) must be documented

* The documentation can be accessed by `help(name)` or `?name`

* For example: `help(seq)`, `help(summary)`, etc.

* A more general (but limited) search can be performed using `help.search("search-string")`

---

# Getting help

* How the help is shown depends on the _interface_ being used

* RStudio has a separate help tab (which also allows searching)
--

* However, before using the help system, you should know a bit about
  how methods are documented

---

# Help for generic functions and methods

* Generic functions and methods are distinct functions

* They often have different help pages

* In fact, many add-on packages define new methods for generics in
  another package

* These are always documented in a separate help page

---

# Help for generic functions and methods

* To get help for the generic function `summary()`, type `help(summary)`

* To get help for the `summary()` method for "matrix" objects, type `help(summary.matrix)`

* To get help for the `summary()` method for "lm" objects, type `help(summary.lm)`

* The first two happen to be the same help page, but the third is different
--

* This is slightly confusing because you are __not__ supposed to call
  `summary.lm()` directly

* More importantly, there may not actually be a `summary()` method for
  all classes 
  
* For example, "list" objects are handled by a _fallback_ method `summary.default()`
--

* The list of available methods are obtained by `methods("summary")`
  as shown earlier
  
* All these should have a corresponding help page

---

# Help for generic functions and methods

- The system we described is called "S3" (short for "S version 3")

- The documentation refers to specific methods implemented using this system as "S3 methods"

- To make things more complicated, there is a different system of
  defining classes and methods
  
- We will skip the details of this system (called "S4") for now


<!-- 

# Other important topics

* How matrices and arrays are implemented in R

	- xtabs() / table() as inverse of as.data.frame.table()
	
	- Vectorization: *apply family

* Formula interface

* Non-standard evaluation

* Replacement functions

* Expressions (as in `replicate(n, rnorm(10))` vs `rep(rnorm(10), n)` )

	- quote(rnorm(10))
	
	- eval(quote(rnorm(10)))

-->

---

class: center middle

# Questions?


