---
layout: remark
title: Non-Standard Evaluation - 1
subtitle: Introductory Computer Programming
author: Deepayan Sarkar
---


# R is not C

* R syntax is superficially similar to C

* This makes it easy to start using R if you are already familiar with C

* However, the internals of how R works is very different (even apart from vectorized arithmetic)

* We will try to go over some of these differences and highlight R "good practices" 


```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
opts_chunk$set(cache = TRUE, cache.path='~/knitr-cache/icp-nseval-1/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
               knitr.table.format = "html",
               fig.width = 15, fig.height = 6, dpi = 96, fig.path='figures/nseval-1-')
options(warnPartialMatchDollar = FALSE, width = 100)
```

---

<!--

# Some important features of R

* Lazy evaluation

* Working with "expressions"

* Non-standard evaluation

* Functions can be arguments of other functions

* The `*apply` family of functions

* Working with matrices

* Basic linear models

-->


# Topics we will cover today

* Lazy evaluation

* Working with "expressions"

* Non-standard evaluation

<!--  

# Example: the probability problem discussed in class

```{r}
probdiv1 <- function(n, nrep)
{
    count <- 0
    for (i in seq_len(nrep)) { # or i in 1:nrep
        s <- sort(sample(n, 2))
        if (s[2] %% s[1] == 0) count <- count + 1
    }
    return (count / nrep)
}
probdiv3 <- function(n)
{
    a <- seq_len(floor(n/2))
    counts <- floor(n / a) - 1
    return (sum(counts) / choose(n, 2))
}
```

-->

---

layout: true

# Lazy evaluation

---

* Arguments passed as function arguments are _not_ evaluated until they are used

* This might seem like a very minor difference

* This feature allows for some interesting behaviour in R

---

```{r}
rchoose1 <- function(u, a, b)
{
    if (u < 0.5) a else b
}
set.seed(20210101) # makes subsequent "random" numbers reproducible
rchoose1(runif(1), sqrt(2), sqrt(-2))
rchoose1(0.25, sqrt(2), sqrt(-2))
rchoose1(0.75, sqrt(2), sqrt(-2))
```

---

* Compare with a similar function written in C++

```{Rcpp}
#include <Rcpp.h>

double cchoose1(double u, double a, double b)
{
    if (u < 0.5) return a;
	else return b;
}

double wsqrt(double x)
{
	if (x < 0) Rcpp::warning("negative input: NaN produced");
	return sqrt(x);
}

// [[Rcpp::export]]
double cchooseSqrt(double u, double a, double b)
{
	return cchoose1(u, wsqrt(a), wsqrt(b));
}
```

---

```{r}
set.seed(20210101)
cchooseSqrt(runif(1), 2, -2)
cchooseSqrt(0.25, 2, -2)
cchooseSqrt(0.75, 2, -2)
```

---

layout: true

# Quoted expressions and evaluation

---

* Remember that R works by evaluating "expressions" typed at the prompt

* Consider the following expression:

```{r}
a < b
```

* This fails because the variables `a` and `b` used in the expression are not defined

---

* It is possible to store an expression without _evaluating_ it in R

* This can be done is several ways; for example

```{r}
e1 <- quote(a < b)
e2 <- expression(a < b)
```

---

* We can explore the nature of these objects in the usual ways

```{r}
class(e1)
length(e1)
str(e1)
```

---

* `expression()` essentially allows a multiple quoted expressions to be stored (as a vector)

```{r}
class(e2)
length(e2)
length(e2[[1]])
str(e2[[1]])
```

---

* Why is this useful?

* Among other things, expressions can be evaluated using the `eval()` function

```{r}
x <- 10
eval(quote(sqrt(x)))
```

---

* Trying to evaluate `e1` again gives the same error because `a` and `b` are not defined

```{r}
e1
eval(e1)
```

--


* However, `eval()` has an optional `envir` argument that can be used to provide missing variables

```{r}
d <- list(a = rnorm(10), b = rexp(10))
eval(e1, envir = d)
```

---

layout: true

# Lazy evaluation and `substitute()`

---

* Lazy evaluation allows a special trick

* An R function can quote an argument without evaluating it

* This is done using the function `substitute()`

```{r}
getExpression <- function(x)
{
    return(substitute(x))
}
getExpression(sqrt(10))
getExpression(a + b)
```

---

* This can be often quite useful

* For example, the following function attaches an attribute storing
  the call to the result of any evaluation

```{r}
withCall <- function(e)
{
    ans <- eval(e)
    attr(ans, "call") <- substitute(e)
    return(ans)
}
withCall(sqrt(10))
withCall(rbind(rnorm(5), rexp(5)))
```

---

* This is how `plot()` can construct nice axis labels 

```{r}
x <- seq(-10, 10, length.out = 201)
plot(x = x, y = x * sin(x), type = "l") 
```

---

* And how `curve()` determines what function to plot

```{r}
curve(x * sin(x), from = -10, to = 10)
```


---

layout: false
class: middle, center

# Questions?


<!--  


```{r}
e1[[1]]
e1[[2]]
e1[[3]]
```


```{r}
e1[[1]] <- quote(rbind)
e1
eval(e1, envir = d)
```

-->
