---
layout: remark
title: Replicate and the Apply Family of Functions - 2
subtitle: Introductory Computer Programming
author: Deepayan Sarkar
mathjax: true
---

<!--  

# Topics 

* Replicating a simulation experiment

* The `*apply()` family of functions

* Functions can be arguments of other functions

* `do.call()`

-->

# The `*apply()` family of functions

* `replicate()` re-evaluates the same expression multiple times

* We often want to something similar:

	- Evaluate the same expression multiple times, but with different arguments
	
* This is what the various `*apply()` functions are intended for

```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
opts_chunk$set(cache = FALSE, cache.path='~/knitr-cache/icp-rapply-2/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
               knitr.table.format = "html",
               fig.width = 15, fig.height = 6, dpi = 96, fig.path='figures/rapply-2-')
options(warnPartialMatchDollar = FALSE, width = 100)
```

---

layout: true

# The `lapply()` function

---

* The most basic apply-type function is `lapply()`

* It takes two mandatory arguments: `lapply(X, FUN)`

	- `X` is a vector object, with `i`-th element given by `X[[i]]`

	- `FUN` is a function
	
* The result is a _list_ object, with the same length as `X`

* The `i`-th element of the result is `FUN(X[[i]])`

---

* Example: 

```{r}
lapply(1:5, sample)
```

* This is essentially a "vectorized" version of the for loop

```{r}
ans <- list()
for (i in 1:5) ans[[i]] <- sample(i) # sample(n) produces a random permutation of 1:n
```

---

* Unlike `replicate()`, the second argument of `lapply()` is not an expression
	
* In other words, `lapply()` does not do non-standard evaluation — whereas `replicate()` does
	
* `lapply()` _always_ returns a list — similar to `simplify = FALSE` in `replicate()`
--

* The somewhat novel feature of `lapply()` is that the `FUN` argument must be a _function_

* This is actually another important difference between R and C: 

	- Functions are not "special"
	
	- They can be assigned to variables
	
	- They can be supplied as arguments to other functions
	
	- They can even be "anonymous", i.e., they don't need to be
      associated with a name
	  
	- In short, they behave just as any other conventional object (vectors, lists, etc.)
	
---

* In fact, it is expected that R users will be able to write short functions when necessary

* Consider our previous example:

```{r,eval=FALSE}
lapply(1:5, sample)
```

* This is conceptually equivalent to 

```{r, eval=FALSE}
ans <- list()
for (i in 1:5)
    ans[[i]] <- sample(i)
```

* But suppose we instead want

```{r, eval=FALSE}
for (i in 1:5)
    ans[[i]] <- sample(10, i, replace = TRUE)
```

---

* `FUN = sample` will no longer serve our purpose, but we can instead define

```{r}
sample_of_size <- function(i)
{
    return(sample(10, i, replace = TRUE))
}
lapply(1:5, FUN = sample_of_size)
```

---

* In fact, there is no need to _name_ the function by storing it in a variable

```{r}
lapply(1:5, FUN = function(i) sample(10, i, replace = TRUE))
```

* This idea is often a big conceptual stumbling block for new R users

---

* A potentially useful feature of `lapply()`:

```{r,eval=FALSE}
lapply(X, FUN, ...) # with further arguments
```

* is equivalent to

```{r, eval=FALSE}
for (i in seq_len(length(X)))
    ans[[i]] <- FUN(X[[i]], ...)
```

---

* For example, this allows

```{r}
str(sample)
lapply(1:5, FUN = sample, size = 20, replace = TRUE)
```

---

* As named arguments match in preference to position, we can even do

```{r}
lapply(1:6, FUN = sample, x = month.abb, replace = TRUE) # sample(x = month.abb, size = i, replace = TRUE)
```

---

* Another example: $N(0, \sigma^2)$ variates with varying $\sigma$

```{r}
str(rnorm)
str(lapply(1:5, FUN = rnorm, n = 1, mean = 0))
```

* But much easier to do directly using vectorization

```{r}
rnorm(5, mean = 0, sd = 1:5)
```

---

* A more complicated example: IQR of 1000 $N(0, \sigma^2)$ variates with varying $\sigma$

```{r}
str(lapply(1:8, function(s) IQR(rnorm(n = 1000, mean = 0, sd = s))))
```

---

layout: true

# `sapply()` : simplifying results to vector / array

---

* `sapply()` is essentially the same as `lapply()`, but simplifies result if possible

```{r}
sapply(1:8, function(s) IQR(rnorm(n = 1000, mean = 0, sd = s)))
```

```{r}
sapply(1:6, function(s) quantile(rnorm(n = 1000, mean = 0, sd = s)))
```

---

layout: true

# `apply()` : Apply function to rows / columns of a matrix

---

* We will talk about matrix operations in R in more detail later

* It is often useful to apply a function on rows or columns of a matrix

```{r}
m1 <- matrix(0, nrow = 1000, ncol = 8)
for (i in 1:8)
    m1[, i] <- rnorm(1000, mean = 0, sd = i)
str(m1)
```

* Note that this is essentially the same as

```{r}
m2 <- sapply(1:8, function(s) rnorm(n = 1000, mean = 0, sd = s))
str(m2)
```

---

* Suppose we want to compute the mean for each column of `m1`

* There are several ways to do this:

	- Loop over columns
	
	- Write this as a matrix operation (will discuss later)
	
	- Use the `colMeans()` or `colSums()` function
	
```{r}
colMeans(m1)
colSums(m1) / nrow(m1)
```

---

* However, `colMeans()` / `colSums()` and `rowMeans()` / `rowSums()` are special

* Other summary functions do not have a built-in row-wise / column-wise versions

* `apply()` allows arbitrary functions to be called per row (`MARGIN=1`) or column (`MARGIN=2`) 

```{r}
apply(m1, 2, mean)
apply(m1, 2, sd)
apply(m1, 2, IQR)
```

---

* `apply()` also allows additional arguments to be specified

```{r}
apply(m1, 2, quantile, probs = c(0.25, 0.75))
```

* As well as unnamed functions

```{r}
apply(m1, 2, function(x) median(abs(x)))
```

---

* Applying a function row-wise

```{r}
apply(m1, 1, function(x) which.max(abs(x)))
```

---

* Applying a function row-wise

```{r,fig.height=4}
wmax <- apply(m1, 1, function(x) which.max(abs(x)))
table(wmax)
barplot(table(wmax))
```

---

* `apply()` can also summarize over margins of higher-dimensional arrays

```{r}
Titanic[,,2,]
```


---

* `apply()` can also summarize over margins of higher-dimensional arrays

```{r}
apply(Titanic[,,2,], MARGIN = 3, sum)
```

* More interestingly, the summary can happen over multiple dimensions

```{r}
apply(Titanic[,,2,], MARGIN = c(2, 3), sum)
```



---

layout: false
class: middle, center

# Questions?

