---
layout: remark
title: Replicate and the Apply Family of Functions - 2
subtitle: Introductory Computer Programming
author: Deepayan Sarkar
mathjax: true
---

<!--  

# Topics 

* Replicating a simulation experiment

* The `*apply()` family of functions

* Functions can be arguments of other functions

* `do.call()`

-->

# The `*apply()` family of functions

* `replicate()` re-evaluates the same expression multiple times

* We often want to something similar:

	- Evaluate the same expression multiple times, but with different arguments
	
* This is what the various `*apply()` functions are intended for

```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
opts_chunk$set(cache = FALSE, cache.path='~/knitr-cache/icp-rapply-2/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
               knitr.table.format = "html",
               fig.width = 15, fig.height = 6, dpi = 96, fig.path='figures/rapply-2-')
options(warnPartialMatchDollar = FALSE, width = 100)
```

---

layout: true

# The `lapply()` function

---

* The most basic apply-type function is `lapply()`

* It takes two mandatory arguments: `lapply(X, FUN)`

	- `X` is a vector object, with `i`-th element given by `X[[i]]`

	- `FUN` is a function
	
* The result is a _list_ object, with the same length as `X`

* The `i`-th element of the result is `FUN(X[[i]])`

---

* Example: 

```{r}
lapply(1:5, sample)
```

* This is essentially a "vectorized" version of the for loop

```{r}
ans <- list()
for (i in 1:5) ans[[i]] <- sample(i) # sample(n) produces a random permutation of 1:n
```

---

* Unlike `replicate()`, the second argument of `lapply()` is not an expression
	
* In other words, `lapply()` does not do non-standard evaluation — whereas `replicate()` does
	
* `lapply()` _always_ returns a list — similar to `simplify = FALSE` in `replicate()`
--

* The somewhat novel feature of `lapply()` is that the `FUN` argument must be a _function_

* This is actually another important difference between R and C: 

	- Functions are not "special"
	
	- They can be assigned to variables
	
	- They can be supplied as arguments to other functions
	
	- They can even be "anonymous", i.e., they don't need to be
      associated with a name
	  
	- In short, they behave just as any other conventional object (vectors, lists, etc.)
	
---

* In fact, it is expected that R users will be able to write short functions when necessary

* Consider our previous example:

```{r,eval=FALSE}
lapply(1:5, sample)
```

* This is conceptually equivalent to 

```{r, eval=FALSE}
ans <- list()
for (i in 1:5)
    ans[[i]] <- sample(i)
```

* But suppose we instead want

```{r, eval=FALSE}
for (i in 1:5)
    ans[[i]] <- sample(10, i, replace = TRUE)
```

---

* `FUN = sample` will no longer serve our purpose, but we can instead define

```{r}
sample_of_size <- function(i)
{
    return(sample(10, i, replace = TRUE))
}
lapply(1:5, FUN = sample_of_size)
```

---

* In fact, there is no need to _name_ the function by storing it in a variable

```{r}
lapply(1:5, FUN = function(i) sample(10, i, replace = TRUE))
```

* This idea is often a big conceptual stumbling block for new R users

---

* A potentially useful feature of `lapply()`:

```{r,eval=FALSE}
ans <- lapply(X, FUN, ...) # with further arguments
```

* is equivalent to

```{r, eval=FALSE}
for (i in seq_len(length(X)))
    ans[[i]] <- FUN(X[[i]], ...)
```

---

* For example, this allows

```{r}
str(sample)
lapply(1:5, FUN = sample, size = 20, replace = TRUE)
```

---

* As named arguments match in preference to position, we can even do

```{r}
lapply(1:6, FUN = sample, x = month.abb, replace = TRUE) # sample(x = month.abb, size = i, replace = TRUE)
```

---

layout: false
class: middle, center

# Questions?

