---
layout: remark
title: Replicate and the Apply Family of Functions - 1
subtitle: Introductory Computer Programming
author: Deepayan Sarkar
mathjax: true
---

<!--  

# Topics 

* Replicating a simulation experiment

* The `*apply()` family of functions

* Functions can be arguments of other functions

* `do.call()`

-->


# Loop alternatives

* R has several general purpose functions that provide alternatives to loops

* We will discuss two of the main approaches today: 

	- `replicate()` for repeating a simulation experiment
	
	- Various `*apply()` functions to loop over varying function arguments 

* These are not necessarily more efficient than loops (unlike true vectorized operations)

* However, they are conceptually simpler (and easier to parallelize)


```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
opts_chunk$set(cache = FALSE, cache.path='~/knitr-cache/icp-rapply-1/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
               knitr.table.format = "html",
               fig.width = 15, fig.height = 6, dpi = 96, fig.path='figures/rapply-1-')
options(warnPartialMatchDollar = FALSE, width = 100)
```

---

layout: true

# Replicating a simulation experiment

---

* Consider this example from our previous assignment:

```{r}
n <- 20
x <- sample(n, 2)
max(x) %% min(x) == 0
```

* To estimate the probability of this event, we want to run it several
  times and compute proportion

---

* We have seen how to do this using a loop

* An alternative approach is to use the `replicate()` function

```{r}
replicate(10, {
    x <- sample(n, 2)
    max(x) %% min(x) == 0
})
```

* `replicate()` takes the _expression_ supplied as its second argument
  and evauates it multiple times

```{r}
s <- replicate(5000, {
    x <- sample(n, 2)
    max(x) %% min(x) == 0
})
sum(s) / length(s) # estimated probability of success
```

---

layout: true

# Return value of `replicate()`

---

* The return value of `replicate()` depends on what the expression returns when evaluated

* In the above example, each evaluation results in a scalar logical

* In that case, the results are aggregated to produce a vector
--

* If the result is always a vector of same length (greater than one), the result is combined into a matrix

```{r}
quantile(rnorm(500), probs = c(0.25, 0.5, 0.75))
replicate(6, quantile(rnorm(500), probs = c(0.25, 0.5, 0.75)))
```

---

* If the result has inconsistent length, the result is a list

```{r}
set.seed(20200101)
replicate(6, unique(sample(10, 5, replace = TRUE)))
```

---

* This may sometimes lead to unanticipated behaviour

```{r}
set.seed(20200101)
replicate(2, unique(sample(10, 5, replace = TRUE)))
```
--

* When output length is known to be variable, it is better to
  explicitly disable simplification

```{r}
set.seed(20200101)
replicate(2, unique(sample(10, 5, replace = TRUE)), simplify = FALSE) # always return list
```

---

layout: false
class: middle, center

# Questions?

