---
layout: remark
title: Replicate and the Apply Family of Functions - 3
subtitle: Introductory Computer Programming
author: Deepayan Sarkar
mathjax: true
---

<!--  

# Topics 

* Replicating a simulation experiment

* The `*apply()` family of functions

* Functions can be arguments of other functions

* `do.call()`

-->


# Recap

* `lapply(X, FUN)` applies the function `FUN` on each element of `X`
	
* Extra arguments to `FUN` can also be supplied (but these remain fixed for each call)
	
* These extra arguments can be named to control which argument of `FUN` varies


```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
opts_chunk$set(cache = FALSE, cache.path='~/knitr-cache/icp-rapply-3/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
               knitr.table.format = "html",
               fig.width = 15, fig.height = 6, dpi = 96, fig.path='figures/rapply-3-')
options(warnPartialMatchDollar = FALSE, width = 100)
```

---

layout: true

# The `lapply()` function

---

* Example: $N(0, \sigma^2)$ variates with varying $\sigma$

```{r}
str(rnorm)
str(lapply(1:5, FUN = rnorm, n = 1, mean = 0))
```
--


* But much easier to do directly using vectorization

```{r}
rnorm(5, mean = 0, sd = 1:5)
```

---

* A more complicated example: IQR of 1000 $N(0, \sigma^2)$ variates with varying $\sigma$

```{r}
str(lapply(1:8, function(s) IQR(rnorm(n = 1000, mean = 0, sd = s))))
```

---

layout: true

# `sapply()` : simplifying results to vector / array

---

* `sapply()` is essentially the same as `lapply()`, but simplifies result if possible

```{r}
sapply(1:8, function(s) IQR(rnorm(n = 1000, mean = 0, sd = s)))
```

```{r}
sapply(1:6, function(s) quantile(rnorm(n = 1000, mean = 0, sd = s)))
```

---

layout: true

# `apply()` : Apply function to rows / columns of a matrix

---

* We will talk about matrix operations in R in more detail later

* It is often useful to apply a function on rows or columns of a matrix

```{r}
m1 <- matrix(0, nrow = 1000, ncol = 8)
for (i in 1:8)
    m1[, i] <- rnorm(1000, mean = 0, sd = i)
str(m1)
```

* Note that this is essentially the same as

```{r}
m2 <- sapply(1:8, function(s) rnorm(n = 1000, mean = 0, sd = s))
str(m2)
```

---

* Suppose we want to compute the mean for each column of `m1`

* There are several ways to do this:

	- Loop over columns
	
	- Write this as a matrix operation (will discuss later)
	
	- Use the `colMeans()` or `colSums()` function
	
```{r}
colMeans(m1)
colSums(m1) / nrow(m1)
```

---

* However, `colMeans()` / `colSums()` and `rowMeans()` / `rowSums()` are special

* Other summary functions do not have a built-in row-wise / column-wise versions

* `apply()` allows arbitrary functions to be called per row (`MARGIN=1`) or column (`MARGIN=2`) 

```{r}
apply(m1, 2, mean)
apply(m1, 2, sd)
apply(m1, 2, IQR)
```

---

* `apply()` also allows additional arguments to be specified

```{r}
apply(m1, 2, quantile, probs = c(0.25, 0.75))
```

* As well as unnamed functions

```{r}
apply(m1, 2, function(x) median(abs(x)))
```

---

* Applying a function row-wise

```{r}
apply(m1, 1, function(x) which.max(abs(x)))
```

---

* Applying a function row-wise

```{r,fig.height=4}
wmax <- apply(m1, 1, function(x) which.max(abs(x)))
table(wmax)
barplot(table(wmax))
```

---

* `apply()` can also summarize over margins of higher-dimensional arrays

```{r}
Titanic[,,2,]
```


---

* `apply()` can also summarize over margins of higher-dimensional arrays

```{r}
apply(Titanic[,,2,], MARGIN = 3, sum)
```

* More interestingly, the summary can happen over multiple dimensions

```{r}
apply(Titanic[,,2,], MARGIN = c(2, 3), sum)
```

---

layout: true

# `tapply()` : Apply function over "ragged" array

---

* Similar in spirit to `apply()`, but meant for groups of unequal size

* General form: `tapply(X, INDEX, FUN, ..., simplify = TRUE)`

	* `X` is split according to grouping variable(s) `INDEX`
	
	* `FUN` applied to each subset

---

* A simple example we have seen before

```{r}
with(mtcars, tapply(mpg, am, FUN = mean))
```

---

* `FUN` can be more complicated

```{r}
with(mtcars, tapply(mpg, am, FUN = summary))
```

---

* Output of different size leads to list output (controlled by `simplify` as usual)

```{r}
with(mtcars, tapply(mpg, am, FUN = unique))
```

---

* Grouping by multiple variables is possible

* The default result is an array with groups as margins

```{r}
with(mtcars, tapply(mpg, list(am, gear), FUN = mean))
```

* Some combinations are empty in this case

```{r}
with(mtcars, tapply(mpg, list(am, gear), FUN = length))
```

---

* `tapply()` essentially calls `split()` followed by `lapply()`

* `split()` is often very useful to divide data into subsets for further manipulation

.scrollable400[

```{r}
str(split(mtcars, mtcars$am))
```

]


---

layout: false

# What else?

```{r}
apropos("apply") # objects with 'apply' in their names
```

* Will not discuss in detail:

	* `vapply()`: Safer version of `sapply()` with pre-specified return type

	* `mapply()`: Vectorized apply, where multiple arguments can be varied

	* `eapply()`: Apply over elements in an environment, which we can ignore for now

	* `rapply()`: Recursive apply (over complex lists), which we can also ignore for now

* Others are even more specialized

---

layout: false
class: middle, center

# Questions?

