---
layout: remark
title: Some Miscellaneous Topics
subtitle: Introductory Computer Programming
author: Deepayan Sarkar
mathjax: true
---


# Some miscellaneous topics

* Matrices and arrays: construction and operations

* Replacement functions

* `do.call()`


```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
opts_chunk$set(cache = TRUE, cache.path='~/knitr-cache/icp-misc-1/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
               knitr.table.format = "html",
               fig.width = 15, fig.height = 6, dpi = 96, fig.path='figures/misc-1-')
library(lattice)
lattice.options(default.args = list(as.table = TRUE), 
                default.theme = list(grid.pars = list(cex = 1.5)))
options(warnPartialMatchDollar = FALSE, width = 100)
```

---

layout: true

# Matrix construction

---

* `matrix(data, nrow, ncol)` : data + dimension

```{r}
matrix(month.abb, 3, 4)
matrix(month.abb, 3, 4, byrow = TRUE)
```

* `array(data, dim)` similarly allows construction of general arrays

---

* `rbind(...)` : construct matrix by combining rows

* `cbind(...)` : construct matrix by combining columns

.scrollable400[

```{r}
X <- with(mtcars, cbind(Int = 1, am = am, wt = wt)) # design matrix
X
```

]

---

layout: true

# Matrix operations

---

* `t(x)` : transpose of matrix

* `aperm(x, perm)` : general array permutation

* `a * b` : element-wise multiplication

* `a %*% b` : matrix multiplication

* `crossprod(x)` : More efficient version of `t(x) %*% x`

* `crossprod(x, y)` : More efficient version of `t(x) %*% y`

* `tcrossprod(x, y)`: Similar version of `x %*% t(y)`

---

```{r}
t(X) %*% X
crossprod(X)
```

---

* `solve(A, b)` : solve system of linear equations $Ax = b$

```{r}
y <- mtcars$mpg
solve(crossprod(X), crossprod(X, y))
```

* Compare with

```{r}
coef(lm(mpg ~ 1 + am + wt, data = mtcars))
```

---

* `lm()` is more numerically efficient (it does not actually compute `crossprod(X)`)

* It also handles factor (categorical) variables through its formula interface

* The `model.matrix()` function computes just the design matrix without fitting the model

---

layout: true

# Replacement functions

---

* We have seen several examples of the form:

```r
x[i] <- value
attr(x, "name") <- value
colnames(x) <- value
```

* All these calls have a "complex" expression on the LHS of the assignment

* These are known as _replacement functions_
--

* These are very natural once you get used to them

* However, they are somewhat unusual, and we should be aware of why they are needed

* The main reason is that R functions are not allowed to _modify_ their arguments

---

* Example: setting names

```{r}
d <- data.frame(1, mtcars$wt, mtcars$am)
names(d)
```

* Suppose we want to change the names to `"Int", "wt", "am"`

* In other languages, we would have something like

```r
setNames(d, c("Int", "wt", "am"))
```

---

* Unfortunately, such a function _cannot_ modify `d`

* The best it can do is to return a modified version of `d`:

```r
d <- setNames(d, c("Int", "wt", "am"))
```

* This is similar to how `transform()`, `dplyr::mutate()`, etc., work

---

* Replacement functions provide a more "natural" alternative

```{r}
str(d)
names(d) <- c("Int", "wt", "am")
str(d)
```

* It is easy to write your own replacement functions

* For now, those details are not important


---

layout: true

# The `do.call()` function to call a function

---

* It is easy to manipulate the language itself in R (e.g., `quote()`)

* Such manipulation is usually only needed for advanced programming, not routine use

* But one particular tool is very useful in practice: `do.call()`


---

* Example

```{r}
am.summary <- with(mtcars, tapply(mpg, am, FUN = summary))
am.summary
```

* Suppose we want to combine these into a matrix:

```{r}
cbind(am.summary[[1]], am.summary[[2]])
```

---

```{r}
gear.summary <- with(mtcars, tapply(mpg, gear, FUN = summary)) # different variable
gear.summary
cbind(gear.summary[[1]], gear.summary[[2]], gear.summary[[3]]) # three arguments instead of two 
```

---

* Alternative: `do.call(what, args)`

	- `what` is a function
	
	- `args` is a _list_ of arguments
	
	- Calls the function `what` with arguments given in `args`

* The following are equivalent:

```r
FUN(val1, val2, arg3 = val3)
```

* and

```r
do.call(FUN, args = list(val1, val2, arg3 = val3))
```

---

* So we can rewrite the previous `cbind()` calls as

```{r}
do.call(cbind, am.summary)
do.call(cbind, gear.summary)
```


---

layout: false
class: middle, center

# Questions?

