---
layout: remark
title: 'Algorithms: Miscellanous Topics'
subtitle: Introductory Computer Programming
author: Deepayan Sarkar
mathjax: true
---



# General lower bound for comparison-based sort

<div>
$$
\newcommand{\sub}[2]{ {#1}_{#2} }
\newcommand{\sumlimits}[2]{ \sum\limits_{#1}^{#2} }
$$
</div>

* For example, this is what happens when we do insertion sort on three
  elements $\sub{a}{1}, \sub{a}{2}, \sub{a}{3}$

* Here, $i \leq j$ denotes the act of comparing $\sub{a}{i}$ and $\sub{a}{j}$

```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
opts_chunk$set(## cache = TRUE, cache.path='~/knitr-cache/icp-algo-misc-1/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
               knitr.table.format = "html",
               fig.width = 15, fig.height = 6, dpi = 96, fig.path='figures/algo-misc-1-')
options(warnPartialMatchDollar = FALSE, width = 100)
```


```{r, echo=FALSE,fig.width=19,fig.height=7}
heap_graph <- function(size, values = NULL)
{
	## need to calculate x and y coordinates of each node
	h <- floor(log2(size))
	stopifnot(size > 1, h > 0)
	xpos <- function(level = 1)
		## x-position of nodes at this level
	{
		## level counted from bottom (lowest layer is level=1)
		## rule: gap between nodes at level l is l
		##       2^(h - l + 1) nodes in total, average position is 0
		x <- seq(1, by = level, length.out = 2^(h-level+1))
		data.frame(x = x - mean(x), y = level)
	}
	d <- do.call(rbind, lapply((h+1):1, xpos))[1:size, ]
	if (!is.null(values)) d <- cbind(d, values = values[1:size])
    class(d) <- c("heap_graph", class(d))
	d
}
plotBinaryTree <- function(d, pch = 16, ..., gp.text = list(cex = 2), newpage = TRUE)
{
	require(grid, quietly = TRUE)
	if (newpage) grid.newpage()
	pushViewport(viewport(xscale = extendrange(d$x, f = 0.15),
	                      yscale = extendrange(d$y, f = 0.15)))
	keep <- !is.na(d$values)
	## add lines joining child to parent
	i <- seq_len(nrow(d))[keep]
	i <- i[-1] # exclude root
	parent <- floor(i/2)
	with(d, grid.segments(x[i], y[i], x[parent], y[parent], 
	                      default.units = "native"))
	with(d, grid.text(rep(c("TRUE", "FALSE"), 5),
                      0.5 * (x[i] + x[parent]), 0.5 * (y[i] + y[parent]), 
	                  default.units = "native", hjust = c(1.1, -0.1), vjust = -0.1))
	## add the points (note: must fill to hide lines)
    with(d, grid.points(x[keep], y[keep], pch = pch, gp = gpar(...)))
	## add text if any
	if (!is.null(d$values)) 
	    with(d, grid.text(parse(text = values[keep]), x[keep], y[keep], 
		                  gp = do.call(gpar, gp.text), default.units = "native"))
	popViewport()
}
## fake binary tree by putting NA values for missing nodes in a heap
#h <- heap_graph(15, values = c("1 ≤ 2", 
#                               "2 ≤ 3", "1 ≤ 3",
#							   "( 1, 2, 3 )", "1 ≤ 3", "( 2, 1, 3 )", "2 ≤ 3", 
#							   NA, NA, "( 1, 3, 2 )", "( 3, 1 ,2 )", NA, NA, "( 2, 3, 1)", "( 3, 2, 1 )"))
h <- heap_graph(15, values = c("1 <= 2", 
                               "2 <= 3", "1 <= 3",
							   "'( 1, 2, 3 )'", "1 <= 3", "'( 2, 1, 3 )'", "2 <= 3", 
							   NA, NA, "'( 1, 3, 2 )'", "'( 3, 1 ,2 )'", NA, NA, "'( 2, 3, 1)'", "'( 3, 2, 1 )'"))
cols <- rep("black", 11)
cols[c(4, 6, 8:11)] <- "transparent"
textCols <- ifelse(cols == "black", "black", "royalblue")
plotBinaryTree(h, cex = 10, pch = 21, fill = "white", col = cols, 
               gp.text = gpar(col = textCols, cex = 2.2))
```



---

# General lower bound for comparison-based sort

* Generally, this decision tree must be a _binary_ tree (two outcomes
  of each comparison)

* It must have at least $n!$ leaf nodes (one or more for each possible permutation)

* Comparisons needed to reach a particular leaf: length of the path
  from the root node
--

* The worst case number of comparisons is the height of the binary
  tree (longest path)

* A binary tree of height $h$ can have at most $2^h$ leaf nodes
--

* A binary tree with at least $n!$ leaf nodes must have height $h \geq \log_2 n!$

* Using [Stirling's
  approximation](https://en.wikipedia.org/wiki/Stirling%27s_approximation)
  $\log n! = n \log n - n + O(\log n)$,

$$
h \geq \log_2(n!) / \log_2(2) = \Theta(n \log n)
$$ 


---

# Linear time sorting

* Sorting can be done in linear time in some special cases

* As shown above, they cannot be comparison-based algorithms

* Usually, these algorithms put restrictions on possible values in the input array
---

* Example: Input consists of letters (lexical / dictionary sorting)

* Common methods:

	* Counting sort
	
	* Radix sort
--

* Details left for a second semester project


---

layout: true

# Randomly permuting arrays

---

* A common requirement in randomized algorithms is to find a random
  permutation of an input array

* One option: assign random key values to each element, then sort the
  elements according to these keys
  
.algorithm[
.name[`PERMUTE-BY-SORTING(A)`]
n = length(A)
let P[1, ..., n] be a new array
__for__ (i = 1, ..., n) {
    P[i] = RANDOM(1, M)
}
sort A, using P as sort keys
]

* Here $M$ should large enough that the possibility of keys being
  duplicated is small
--

* Exercise: 

	- Assume that all key values are distinct 
	- Show that `PERMUTE-BY-SORTING` produces a uniform random permutation of the input


---

* The runtime for `PERMUTE-BY-SORTING` will be $\Omega(n \log_2 n)$ if we
  use a comparison sort

* A better method for generating a random permutation is to permute the
  given array in place
--
  
* The following algorithm does so in $\Theta(n)$ time

.algorithm[
.name[`RANDOMIZE-IN-PLACE(A)`]
n = length(A)
__for__ (i = 1, ..., n) {
    swap(A[i], A[ RANDOM(i, n) ])
}
]

* In the $i$th iteration, $A[i]$ is chosen randomly from among $A[i],
  A[i+1], ..., A[n]$

* Subsequent to the ith iteration, $A[i]$ is never altered.

---

* Procedure `RANDOMIZE-IN-PLACE` computes a uniform random permutation

* We prove this using the following loop invariant

	> Just prior to the $i$th iteration of the for loop, for each
	> possible $(i-1)$-permutation of the $n$ elements, the subarray
	> $A[1,...,i-1]$ contains this $(i-1)$-permutation with
	> probability $(n-i+1)!/n!$.\
--	

### Initialization

* Holds trivially ($i-1=0$)

* If this is not convincing, take (just before) $i=2$ to be the
  initial step


---

### Maintenance 

* Assume true upto $i=1,...,k$

* Consider what happens just before $i=(k+1)$th iteration (i.e., just after $k$th iteration)

* Let $(\sub{X}1, \sub{X}2, ..., \sub{X}k)$ be the random variable denoting the observed
permutation

*  For any specific $k$-permutation $(\sub{x}1, \sub{x}2, ..., \sub{x}k)$,

<div>
\begin{eqnarray*}
  P(\sub{X}1 = \sub{x}1, \sub{X}2 = \sub{x}2, ..., \sub{X}k = \sub{x}k) 
  &=& P(\sub{X}k = \sub{x}k | \sub{X}1 = \sub{x}1, \sub{X}2 = \sub{x}2, ..., \sub{X}{k-1} = \sub{x}{k-1}) \\
  & & \times  P(\sub{X}1 = \sub{x}1, \sub{X}2 = \sub{x}2, ..., \sub{X}{k-1} = \sub{x}{k-1}) \\
  &=& \frac{1}{n-k+1} \times \frac{(n-k+1)!}{n!} = \frac{(n-k)!}{n!}
\end{eqnarray*}
</div>
--

### Termination

* $i=n+1$, so each $n$-permutation is observed with probability $1/n!$


---

layout: false

# Further topics

* We will not discuss analysis of algorithms further

* If you are interested, an excellent book on this topic is\
  <a
  href="https://www.amazon.in/Introduction-Algorithms-Eastern-Economy-Thomas/dp/8120340078/"
  >Introduction to Algorithms</a> by Cormen, Leiserson, Rivest and Stein



---

layout: false
class: middle, center

# Questions?

